Electrostatics of Salt-Dependent Reentrant Phase Behaviors Highlights Diverse Roles of ATP in Biomolecular Condensates

Version 1

Abstract

Abstract

Liquid-liquid phase separation (LLPS) involving intrinsically disordered protein regions (IDRs) is a major physical mechanism for biological membraneless compartmentalization. The multifaceted electrostatic effects in these biomolecular condensates are exemplified here by experimental and theoretical investigations of the different salt- and ATP-dependent LLPSs of an IDR of messenger RNA-regulating protein Caprin1 and its phosphorylated variant pY-Caprin1, exhibiting, e.g., reentrant behaviors in some instances but not others. Experimental data are rationalized by physical modeling using analytical theory, molecular dynamics, and polymer field-theoretic simulations, indicating in general that interchain salt bridges enhance LLPS of polyelectrolytes such as Caprin1 and that the high valency of ATP-magnesium is a significant factor for its colocalization with the condensed phases, as similar trends are observed for several other IDRs. Our findings underscore the role of biomolecular condensates in modulating ion concentrations and its functional ramifications.

Introduction

Broad-based recent efforts have uncovered many intriguing features of biomolecular condensates, revealing and suggesting myriad known and potential biological functions [1–3]. These assemblies are underpinned substantially, though not exclusively, by liquid-liquid phase separation (LLPS) of intrinsically disordered regions (IDRs) as well as folded domains of proteins and nucleic acids [4, 5], while more complex equilibrium and non-equilibrium mechanisms also contribute [6–14].

Electrostatics is important for IDR LLPS [15, 16], which is often also facilitated by π-related interactions [17, 18], hydrophobicity, hydrogen bonding [19, 20], and is modulated by temperature [10, 21], hydrostatic pressure [22], osmolytes [4], RNA [23–26], salt, pH [27], and post-translational modifications (PTMs) [1, 28, 29]. Multivalency underlies many aspects of IDR properties [30–34]. Here, we focus primarily on how PTM- and salt-modulated multivalent charge-charge interactions might alter IDR condensate behaviors and their possible functional ramifications. In general, electrostatic effects on IDR LLPS [15, 25, 27, 35] are dependent upon their sequence charge patterns [36–40]. Intriguingly, some IDRs undergo reentrant phase separation [4] or dissolution [41] when temperature, pressure [4], salt [42, 43], RNA [41, 44], or concentrations of small molecules such as heparin [45] is varied. Reentrance, especially when induced by salt and RNA, suggest a subtle interplay between multivalent sequence-specific charge-charge interactions and hydrophobic, non-ionic [42, 43], cation-π [41, 44], or π-π interactions.

An important modulator of biomolecular LLPS is adenosine triphosphate (ATP). As energy currency, ATP hydrolysis is utilized to synthesize or break chemical bonds and drive transport to regulate “active liquid” properties such as concentration gradients and droplet sizes [8, 46]. Examples include ATP-driven assembly of stress granules [47], splitting of bacterial biomolecular condensates [48], and destabilization of nucleolar aggregates [49]. ATP can also influence biomolecular LLPS without hydrolysis, akin to other LLPS promotors or suppressors [50] that are effectively ligands of the condensate scaffold [51], or through ATP’s effect on lowering free [Mg2+] [52]. Notably, as an amphiphilic hydrotrope [53] with intracellular concentrations much higher than that required for an energy source, ATP is also seen to afford an important function independent of hydrolysis by solubilizing proteins, preventing LLPS and destabilizing aggregates, as exemplified by measurements on several proteins including fused in sarcoma (FUS) [54].

Subsequent investigations indicate, however, that hydrolysis-independent [ATP] effects on biomolecular LLPS are neither invariably monotonic for a given system nor universal across different systems. For instance, ATP promotes, not suppresses, LLPS of an IgG1 antibody [55] and of basic IDPs [56], and ATP enhances LLPS of full-length and the C-terminal domain (CTD) of FUS at low [ATP] but prevents LLPS at high [ATP] [57]. The latter reentrant behavior has been surmised to arise from ATP binding bivalently [57, 58] or trivalently [59] to charged residues arginine (R) or lysine (K) by a combination of cation-π and electrostatic interactions, an effect also seen in the ATP-mediated LLPS of basic IDPs [56]. A similar scenario was invoked for the reentrant phase behavior of transactive response DNA-binding protein of 43 kDa (TDP-43) [60].

While n-related interactions are important for biomolecular LLPS in general [17, 18] and their interplay with electrostatics likely underlies reentrant biomolecular phase behaviors modulated by RNA [41, 44] or simple salts [42], the degree to which electrostatics alone can rationalize hydrolysis-independent ATP-modulated biomolecular phase reentrance has not been sufficiently appreciated. This question deserves attention. For instance, the suppression of cold-inducible RNA-binding protein condensation by ATP has been suggested to be electrostatically driven [61]. The aforementioned ATP-modulated reentrant phase behavior of FUS [57, 58] is reminiscent of the 236-residue N-terminal IDR of DEAD-box RNA helicase Ddx4’s lack of LLPS at low [NaCl] (< 15–20 mM), LLPS at higher [NaCl] [62] and decreasing LLPS propensity when [NaCl] is further increased [15, 16]. Indeed, the finding that FUS CTD (net charge per residue (NCPR) = 15/156 = 0.096) exhibits ATP-dependent reentrant phase behaviors while the N-terminal domain (NCPR = 3/267 = 0.011) does not [58] is consistent with electrostatics-based theory for the difference in salt-dependent LLPS of polyelectrolytes and polyampholytes [62].

With this in mind, we seek to delineate the degree to which theories focusing primarily on electrostatics can rationalize experimental ATP-related LLPS data on the 103-residue C-terminal IDR of human cytoplasmic activation/proliferation-associated protein-1 (Caprin1). Full-length Caprin1 (709 amino acid residues) is a ubiquitously expressed phosphoprotein that regulates stress [63–66] and neuronal [67] granules, is necessary for normal cellular proliferation [68, 69], and may be essential for long-term memory [70]. Caprin1 dysfunction leads to multiple diseases such as nasopharyngeal carcinoma [71] as well as language impairment and autism spectrum disorder [72], via, e.g., Caprin1’s modulation of the function of the fragile X mental retardation protein (FMRP) [24, 28, 67]. The C-terminal 607–709 Caprin1 IDR, referred to simply as Caprin1 below, is biophysically and functionally significant: It is sufficient for LLPS in vitro [28], important for assembling stress granules in the cell [63, 64], and has a substantial body of experiments [28, 73–75] for comparison with theory. Since tyrosine phosphorylations of Caprin1 in vivo [76] may regulate translation in neurons [28], the Caprin1 system is also useful for gaining insights into phosphorregulation of biomolecular condensates [77–79].

Recent advances in theory and computation enable modeling of sequence-specific IDR LLPS [16, 78, 80–86]. Among the approaches, explicit-chain simulation affords more realistic geometric and energetic representations while analytical theory offers advantages in numerical tractability [87]. The analytical rG-RPA formulation [62], which synthesizes Kuhn-lengh renormalization (renormalized Gaussian, rG) and random phase approximation (RPA) [16] to treat both high-net-charge polyelectrolytes and essentially net-neutral polyampholytes [62], is particularly well suited for Caprin1 and its phosphorylated variant. We hereby leverage a methodological combination of rG-RPA [62], field-theoretic simulation [81, 88], and coarse-grained explicit-chain molecular dynamics (MD) [78, 84] to elucidate the effects of salt, phosphorylation, and ATP on Caprin1 LLPS.

Results

Physical theories of Caprin1 and phosphorylated Caprin1 LLPSs as those of polyelectrolytes and polyampholytes

The 103-residue Caprin1 is a highly charged IDR with 19 charged residues (Fig. 1a and SI Appendix, Fig. S1): 15 R, 1 K, and 3 aspartic acids (D); fraction of charged residues = 19/103 = 0.184 and NCPR = 13/103 = 0.126. With a substantial positive net charge, Caprin1’s phase behaviors are markedly different from those of polyampholytic IDRs with nearly zero net charge such as Ddx4 to which early sequence-specific LLPS theories were targeted [15, 16]. Instead, Caprin1 behaves like chemically synthesized polyelectrolytes [89]. In contrast, when most or all of the seven tyrosines (Y) in the Caprin1 IDR are phosphorylated (pY), negative charges are added to produce a near-net-neutral polyampholyte. Mass spectrometry indicates that the experimental sample of phosphorylated Caprin1 consists mainly of a mixture of IDRs with six or seven phosphorylations (SI Appendix, Fig. S2). We refer to this experimental sample as pY-Caprin1 below. For simplicity, we use only the Caprin1 IDR with seven pYs to model the behavior of this highly phosphorylated experimental sample in our theoretical and computational formulations, partly to avoid the combinatoric complexity of sequences with 5 or 6 pYs. Accordingly, since the charge of a pY is ≈ —2 at the experimental pH = 7.4, -14 charges are added to Caprin1 for our model pY-Caprin1, resulting in a polyampholyte with a very small NCPR ≈ -1/103 = -0.00971 (Fig. 1b). Both the experimental pY-Caprin1 (NCPR « ±1/103 = ±0.00971) and model pY-Caprin1 are expected to exhibit phase properties similar to other polyampholytic IDRs.

rG-RPA theory predictions rationalize different salt dependence of Caprin1 and pY-Caprin1 LLPS. (a,b) Vertical lines indicate the sequence positions (horizontal variable) of positively charged residues (blue) and negatively charged residues or phosphorylated tyrosines (red) for (a) Caprin1 and (b) pY-Caprin1. (c,d) rG-RPA coexistence curves (phase diagrams, continuous curves color-coded for the NaCl concentrations indicated) agree reasonably well with experiment (dots, same color code). The grey arrows in (c,d) highlight that when [NaCl] increases, LLPS propensity increases for (c) Caprin1 but decreases for (d) pY-Caprin1.

While sequence-specific RPA has been applied successfully to model electrostatic effects on the LLPSs of various polyampholytic IDRs [10, 16, 84, 90, 91], RPA is less appropriate for polyelectrolytes with large NCPR [92–94] because of its treatment of polymers as ideal Gaussian chains [95]. Traditionally, theories for polyelectrolytes tackle their peculiar conformations by various renormalized blob constructs [89, 92], two-loop polymer field theory [96], modified thermodynamic perturbation theory [97], and renormalized Gaussian fluctuation (RGF) theory [98, 99], among others. As such, these formulations are mostly designed for homopolymers, making it difficult to apply directly to heteropolymeric biopolymers. In order to analyze Caprin1 and pY-Caprin1 LLPSs, we utilize rG-RPA [62], which combines Gaussian chains of effective (renormalized) Kuhn length with the key idea of RGF [37].

Phase properties predicted by rG-RPA theory for Caprin1 and pY-Caprin1 with monovalent counterions and salt are in agreement with experiment

Fig. 1c and d show that the salt- and temperature (T)-dependent phase diagrams predicted by rG-RPA with an augmented Flory-Huggins (FH) mean-field χ(T) = Δh/T — Δs parameter for nonelectrostatic interactions [16, 62] (“rG-RPA+FH” theory in SI Appendix) are in reasonable agreement with experiment using bulk [Caprin1] ≈ 200 μM. The rG-RPA+FH results in Fig. 1c indicate that (i) Caprin1 undergoes LLPS below 20°C with 100 mM NaCl, and that (ii) LLPS propensity, quantified by the upper critical solution temperature (UCST), increases with [NaCl]. These predictions are consistent with experimental data, including the observation that Caprin1 does not phase separate at room temperature without salt, ATP, RNA or other proteins, though Caprin1 LLPS can be triggered by adding wildtype (WT) and phosphorylated FMRP and/or RNA (bulk [Caprin1] ≥ 10 μM) [28], NaCl [73], or ATP (bulk [Caprin1] = 400 μM) [74]. The trend here is also in line with other theories of polyelectrolytes [99]. In contrast, rG-RPA+FH results in Fig. 1d for pY-Caprin1 shows decreasing LLPS propensity with increasing [NaCl], consistent with experimental data and the expected salt dependence of LLPS of nearly net-neutral polyampholytic IDRs such as Ddx4 [16]. Interestingly, the decrease in some of the condensed-phase [pY-Caprin1]s with decreasing T (orange and green symbols for ≤ 20°C in Fig. 1d trending toward slightly lower [pY-Caprin1]) is suggestive of a lower critical solution temperature (LCST)-like [10, 21] reduction of LLPS propensity as temperature approaches ~ 0°C.

Salt-IDR two-dimensional phase diagrams are instrumental for exploring broader phase properties

Fig. 1c and d, though informative, are computed by a restricted rG-RPA that assumes a spatially uniform [Na+]. For a more comprehensive physical picture, we now examine possible differences in salt concentration between the IDR-dilute and condensed phases by applying unrestricted rG-RPA to compute two-dimensional salt-Caprin1/pY-Caprin1 phase diagrams (Fig. 2).

rG-RPA theory rationalizes [NaCl]-modulated reentrant phase behavior of Caprin1. In each salt-protein phase diagram (T = 300 K), tielines (dashed) connect coexisting phases on the boundary (magenta curve) of the cyan-shaded coexistence region. For clarity, zoomed-in views of the grey-shaded part in (a, c, e, g, i, k) are provided by the plots to the right, i.e., (b, d, f, h, j, l), respectively. The solid inclined lines in (g, h, k, l) mark the minimum counterion concentrations required for overall electric neutrality. Results are shown for monovalent cation and anion with Caprin1 (a, b) or pY-Caprin1 (c, d); or monovalent cation and divalent anion with Caprin1 (e-h); or divalent cation and tetravalent anion with Caprin1 (i-l). Cation-modulated reentrant phase behaviors is seen for a wide concentration range for Caprin1 in (a, b) but only a very narrow range of high Caprin1 concentrations in (e, f, i, j).

As stated in Materials and Methods and SI Appendix, here we define “counterions” and “salt ions”, respectively, as the small ions with charges opposite and identical in sign to that of the net charge, Q, of a given polymer. For the Caprin1/NaCl system, since Caprin1’s net charge is positive, Na+ is salt ion and Cl– is counterion. Overall electric neutrality of the system implies that the concentrations (ρ’s) of polymer (ρp), counterions (ρc), and salt ions (ρs) are related by | Q |ρP+zcρc, (1)

where zs and zc are, respectively, the valencies of salt ions and counterions. For Caprin1 and pY-Caprin1, Q = +13 and -1, respectively, and (zs, zc) = (1, 1), (1, 2), and (2, 4) are models for different small-ion species in the system. Specifically, in Fig. 2, we identify the zs = 1 salt ion as Na+ (Fig. 2a-f) and the zc =1 counterion as Cl– (Fig. 2a-d), the zc = 2 counterion as (ATP-Mg)2- (Fig. 2g,h), the zs = 2 salt ion as Mg2+ and the zc = 4 counterion as ATP4- (Fig. 2i-l).

Behavioral trends of rG-RPA-predicted Na+-Caprin1 two-dimensional phase diagrams are consistent with experiment

Notably, Fig. 2a,b (zs = zc = 1) predicts that Caprin1 does not phase separate without Na+ , consistent with experiment, indicating that monovalent counterions alone (Cl– in this case) are insufficient for Caprin1 LLPS. When [Na+] is increased, the system starts to phase separate at a small [Na+] ≤ 0.1 M, with LLPS propensity increasing to a maximum at [Na+] ~1 M before decreasing at higher [Na+], in agreement with experiment (Fig. 3a, blue data points) and consistent with Caprin1 LLPS propensity increasing with [NaCl] from 0.1 to 0.5 M (Fig. 1c). The predicted reentrant dissolution of Caprin1 condensate at high [Na+] in Fig. 2a is consistent with measurement up to [Na+] ≈ 4.6 M indicating a significant decrease in LLPS propensity when [Na+] ≥ 2.5 M (Fig. 3a), though the gradual decreasing trend suggests that complete dissolution of condensed droplets is not likely even when NaCl reaches its saturation concentration of ~ 6 M.

Experimental demonstration of [ATP-Mg]- and [NaCl]-modulated reentrant phase behavior for Caprin1. (a) Turbidity quantified by optical density at 600 nm (OD600, normalized by peak value) to assess Caprin1 LLPS propensity at [Caprin1] = 200 μM [for ATP-Mg dependence (red), bottom scale] or [Caprin1] = 300 μM [for NaCl dependence (blue), top scale], measured at room temperature (~ 23°C). Error bars are one standard deviations of triplicate measurements, which in most cases was smaller than the plotting symbols. The ATP-Mg dependence seen here for 200 μM Caprin1 is similar to the results for 400 μM Caprin1 (Fig. 6C of ref. [74]). (b) Microscopic images of Caprin1 and pY-Caprin1 at varying [ATP-Mg] at room temperature, showing reentrant behavior for Caprin1 but not for pY-Caprin1. Each sample contains 200 μM of either Caprin1 or pY-Caprin1, with 1% of either Caprin1-Cy5 or pY-Caprin1-Cy5 (labeled with Cyanine 5 fluorescent dye) added for visualization, in a 25 mM HEPES buffer at pH 7.4. Scale bars represent 10 μm.

The negative tieline slopes in Fig. 2a,b predict that Na+ is partially excluded from the Caprin1 condensate. This “salt partitioning” is most likely caused by Caprin1’s net positive charge and is consistent with published research on polyelectrolytes with monovalent salt [99–101]. Here, the rG-RPA predicted trend is consistent with our experiment showing significantly reduced [Na+] in the Caprin1-condensed compared to the dilute phase (Table 1), although the larger experimental reduction of [Na+] in the Caprin1 condensed droplet relative to our theoretical prediction remains to be elucidated. In contrast, for the near-neutral, very slightly negative model pY-Caprin1 (Fig. 2c,d), rG-RPA predicts LLPS at [Na+] ≈ 0, and the positive tieline slopes indicate that [Na+] is higher in the condensed than in the dilute phase. Consistent with Fig. 1d, Fig. 2c shows that pY-Caprin1 LLPS propensity always decreases with increasing [Na+].

rG-RPA-predicted salt-IDR two-dimensional phase diagrams underscore effects of counterion valency on LLPS

Interestingly, a different salt dependence of Caprin1 LLPS is predicted when the salt ion remains monovalent but Cl– is replaced by a divalent zc = 2 anion modeling (ATP-Mg)2- under the simplifying assumption that ATP4- and Mg2+ do not dissociate in solution. The corresponding rG-RPA results (Fig. 2e-h) indicate that, with divalent counterions (Fig. 2g,h), Caprin1 can undergo LLPS without the monvalent salt (Na+) ions (LLPS regions extend to [Na+] = 0 in Fig. 2e,f), which is likely casued by the fact that the configurational entropic cost of concentrating counterions in the Caprin1 condensed phase is lesser for divalent than for monovalent counterions.

Other predicted differences between monovalent (Fig. 2a,b) and divalent (Fig. 2e,f) counterions’ impact on Caprin1 LLPS include: (i) The maximum condensed-phase [Caprin1] at low [Na+] is lower with monovalent than with divalent counterions ([Caprin1] ~ 400 mM vs. ~ 750 mM). (ii) The [Na+] at the commencement of reentrance (i.e., at the maximum condensed-phase [Caprin1]) is much higher with monovalent than with divalent counterions ([Na+] ~ 1 M vs. ~ 0.1 M). (iii) [Na+] is depleted in the Caprin1 condensate with both monovalent and divalent counterions when overall [Na+] is high (negative tieline slopes for [Na+]≤ 2 M in Fig. 2a,e). However, for lower overall [Na+], [Na+] is slightly higher in the Caprin1 condensate with divalent but not with monovalent counterions (slightly positive tieline slopes for [Na+]. 2 M in Fig.2e,f). This prediction suggests that under physiological [Na+]=150~170 mM, monovalent positive salt ions such as Na+ can be attracted, somewhat counterintuitively, into biomolecular condensates scaffolded by positively-charged polyelectrolytic IDRs in the presence of divalent counterions.

rG-RPA is consistent with experimental [ATP-Mg]-dependent Caprin1 reentrant phase behaviors

For the zs = 2, zc = 4 case in Fig. 2i-l modeling (ATP-Mg)2- complex dissociating completely in solution into Mg2+ salt ions and ATP4- counterions, rG-RPA predicts Caprin1 LLPS with ATP4- (Fig. 2k,l) in the absence of Mg2+ (the LLPS region includes the horizontal axes in Fig. 2i,j), likely because the configurational entropy loss of tetravalent counterions in the Caprin1 condensate is less than that of divalent and monovalent counterions. Tetravalent counterions also increase the maximum condensed-phase [Caprin1] to ≥1200 mM. At the commencement of reentrance (maximum condensed-phase [Caprin1] in Fig. 2i,j), [Mg2+] ~400 mM, which is intermediate between the corresponding [Na+] ~ 1.0 and 0.1 M, respectively, for monovalent and divalent counterions with (zs, zc) = (1, 2) and (1, 1). All tieline slopes for Mg2+ and ATP4- in Fig. 2i-l are significantly positive, except in an extremely high-salt region with [Mg2+]> 8M, indicating that [ATP-Mg] is almost always substantially enhanced in the Caprin1 condensate. Despite the tendency for polymer field theories to overestimate LLPS propensity and condensed-phase concentrations quantitatively because they do not account for ion condensation [98]—which can be severe for small ions with more than ±1 charge valencies, our rG-RPA-predicted semi-quantitative trends are consistent with experiments indicating [ATP-Mg]-dependent reentrant phase behavior of Caprin1 (Fig. 3a, red data points, and Fig. 3b) and that [Mg2+] as well as [ATP4-] are significantly enhanced in the Caprin1 condensate by a factor of ~ 5–60 for overall [ATP-Mg] = 3–30 mM (Table 2).

Coarse-grained MD with explicit small ions is useful for investigating subtle salt dependence in biomolecular LLPS

To gain deeper insights, we extend the widely-utilized coarse-grained explicit-chain MD model for biomolecular condensates [78, 84, 102] to include explicit small cations and anions (Materials and Methods). For computational efficiency, we neglect solvation effects that can arise from the directional hydrogen bonds among water molecules (see, e.g., ref. [103]) by treating other aspects of the aqueous solvent implicitly as in most, though not all [87, 91] applications of the methodology [78]. Several coarse-grained interaction schemes were used in recent MD simulations of biomolecular LLPS [78, 84, 86, 104–107]. Since we are primarily interested in general principles rather than quantitative details of the phase behaviors of Caprin1 and its ariginine-to-lysine (RtoK) mutants, here we adopt the Kim-Hummer (KH) energies for pairwise amino acid interactions derived from contact statistics of folded protein structures [78], which can largely capture the experimental effects of R vs K on LLPS [84].

Explicit-ion MD rationalizes experimentally observed [NaCl]-dependent Caprin1 reentrant phase behaviors and depletion of Na+ in Caprin1 condensate

Consistent with experiment (Fig. 3) and rG-RPA (Fig. 2a-d), explicit-ion coarse-grained MD shows [NaCl]-dependent reentrant phase behavior for Caprin1 but not for pY-Caprin1 (non-monotonic and monotonic trends indicated, respectively, by the grey arrows in Fig. 4a,b). In other words, the critical temperature Tcr, which is defined as the maximum temperature (UCST) of a given phase diagram (binodal, or coexistence curve), increases then decreases with addition of NaCl for Caprin1 but Tcr always decreases with increasing [NaCl] for pY-Caprin1. Moreover, consistent with the rG-RPA-predicted tielines in Fig. 2a-d (negative slopes for Caprin1 and positive slopes for pY-Caprin1), Fig. 4c shows that Na+ is slightly depleted in the Caprin1 condensed droplet, exhibiting the same trend as that in experiment (Fig. 3a, blue data points; and Table 1) but is enhanced in the pY-Caprin1 droplet (Fig. 4d). Because model temperatures in Fig. 4a,b and subsequent MD results are given in units of the MD-simulated Tcr of WT Caprin1 at [NaCl] = 0 (denoted as Tcr0 here), the Tcrs of systems with higher or lower LLPS propensities than WT Caprin1 at zero [NaCl] is characterized, respectively, by Tcr/Tcr0>1 or < 1.

Explicit-ion coarse-grained MD rationalizes [NaCl]-modulated reentrant behavior for Caprin1 and lack thereof for pY-Caprin1. (a) Simulated phase diagrams (binodal curves) of Caprin1 at different temperatures plotted in units of Tcr0 (see text). Symbols are simulated data points. Continuous curves are guides for the eye. Grey arrow indicates variation in [NaCl]. (b) Same as (a) but for pY-Caprin1. (c) A snapshot showing phase equilibrium between dilute and condensed phases of Caprin1 (brown chains) immersed in Na+ (blue) and Cl– (red) ions simulated at [NaCl] = 480 mM. (d) A similar snapshot for pY-Caprin1. (e, f) Density profiles, ρ(z), of Na+, Cl–, and (e) Caprin1 or (f) pY-Caprin1 along the elongated dimension z of the simulation box showing variations of Na+ and Cl– concentrations between the protein-dilute phase (low ρ for protein) and protein-condensed phase (high ρ for protein) at the simulation temperatures indicated.

Fig. 4c shows that [Cl–] is enhanced while [Na+] is depleted in the Caprin1 droplet. This phenomenon is further illustrated in Fig. 5a-d by comparing the entire simulation box (Fig. 5a) with individual distributions of the Caprin1 IDR (Fig. 5b), Na+ (Fig. 5c), and Cl– (Fig. 5d). A similar trend, also attributed to charge effects, was observed in explicit-water, explicit-ion MD simulations in the presence of a preformed condensate of the N-terminal RGG domain of LAF-1 with a positive net charge [108]. For Caprin1, Fig. 5e,f suggests that, as counterion, Cl– can coordinate two positively charged R residues and thereby stabilize indirect counterion-bridged interchain contacts among polycationic Caprin1 molecules to promote LLPS, consistent with an early lattice-model analysis of generic polyelectrolytes [94].

Counterions can stabilize Caprin1 condensed phase by favorable bridging interactions. (a) Snapshot from explicit-ion coarse-grained MD under LLPS conditions for Caprin1, showing the spatial distributions of Caprin1, Na+, and Cl– (as in Fig. 4c). The three components of the same snapshot are also shown separately in (b) Caprin1, (c) Na+, and (d) Cl–. (e) A zoomed-in view of the condensed droplet [corresponding to the green box in (a)], now with a black background and a different color scheme. (f) A further zoomed-in view of the part enclosed by the green box in (e) focusing on two interacting Caprin1 chains. A Cl– ion (pink bead indicated by the arrow) is seen interacting favorably with two arginines residues (blue beads) on the two Caprin1 chains (whose uncharged residues are colored differently by yellow or orange, lysine and aspartic acids in both chains are depicted, respectively, in magenta and red).

Explicit-ion MD rationalizes [NaCl]-dependent phase properties of argininelysine mutants of Caprin1

We apply our MD methodology also to four RtoK Caprin1 variants, termed 15Rto15K, 4Rto4KN, 4Rto4KM, and 4Rto4KC (SI Appendix, Fig. S1), which involve 15 or 4 RtoK substitutions [73]. The simulated phase diagrams in Fig. 6 exhibit reentrant phase behaviors for all three 4Rto4K variants. While these results are consistent with experiments showing LLPS of these 4Rto4K variants commencing at different nonzero [NaCl]s [73], the simulated reentrant dissolution is not observed experimentally, probably because the actual [NaCl] needed is beyond the experimentally investigated or physically possible range of salt concentration. Simulated reentrant phase behaviors are also seen for 15Rto15K; but as will be explained below, its much lower simulated UCST is consistent with no experimental LLPS for this variant [73]. Since our main focus here is on general physical principles, we do not attempt to fine-tune the MD parameters for a quantitative match between simulation and experiment. Experimentally, only WT exhibits a clear trend toward reentrant dissolution of condensed droplets (with a LLPS propensity plateau at [NaCl] ≈ 1.55–2.5 M, Fig. 3a, blue data points), whereas the LLPS of 4Rto4KM and 4Rto4KC commences at [NaCl] ~ 1.3 M, LLPS propensity then increases with [NaCl] (a trend consistent with the MD-predicted increasing LLPS propensity at low [NaCl]s in Fig. 6b,c), but no sign of reentrant dissolution is seen up to the maximum [NaCl] = 2 M investigated experimentally for the RtoK variants (Fig. 9B of ref. [73]). In contrast, the MD phase diagrams in Fig. 6 show a maximum LLPS propensity (highest Tcr) at [NaCl] ≈ 0.5 M. This qualitative agreement with quantitative mismatch suggests that real Caprin1 LLPS is somewhat less sensitive to small monovalent ions than that stipulated by the present MD model. This question should be tackled in future studies by considering, for example, alternate pairwise amino acid interaction energies [78, 84, 86, 104–107] and their temperature dependence [4, 21].

Explicit-ion coarse-grained MD rationalizes [NaCl]-modulated phase behavior for RtoK variants of Caprin1. Four variants studied experimentally [73] are simulated: (a) 15Rto15K, in which 15 R’s in the WT Caprin1 IDR are substituted by K, (b) 4RtoKN, (c) 4RtoKM, and (d) 4RtoKC, in which 4 R’s are substituted by K in the (b) N-terminal, (c) middle, and (d) C-terminal regions, respectively. Top panels show positions of the R (dark blue) and K (cyan) along the Caprin1 IDR sequence. Lower panels are phase diagrams in the same style as Fig. 4. The phase diagrams for WT Caprin1 from Fig. 4a are included as continuous curves with no data points in (a) for comparison.

Limitations notwithstanding, the MD-simulated trend agree largely with experiment. Predicted LLPS propensities quantified by the Tcrs in Fig. 6 follow the rank order of WT > 4Rto4KM > 4RtoKN ≈ 4Rto4KC > 15Rto15K, which is essentially identical to that measured experimentally, viz., WT > 4Rto4KM > 4RtoKC > 4Rto4KN > 15Rto15K (Fig. 9B of ref. [73]). In comparing theoretical and experimental LLPS, a low theoretical Tcr can practically mean no experimental LLPS when the theoretical Tcr is below the freezing temperature of the real system [16, 109]. Fig. 6a shows that even the highest Tcr for 15Rto15K (at model [NaCl] = 480 mM) is essentially at the same level as Tcr0 for WT at [NaCl] = 0 ( Tcr/Tcr0≈1 ). This MD prediction is consistent with the combined experimental observations of no LLPS for 15Rto15K up to at least [NaCl] = 2 M and no LLPS for WT Caprin1 at [NaCl] = 0 (Fig. 9B,C of ref. [73]).

Field-theoretic simulation is an efficient tool for studying multiple-component phase properties

We next turn to modeling of Caprin1 or pY-Caprin1 LLPS modulated by both ATP-Mg and NaCl. Because tackling such many-component LLPS systems using rG-RPA or explicit-ion MD is numerically challenging, here we adopt the complementary field-theoretic simulation (FTS) approach [110] outlined in Materials and Methods for this aspect of our investigation. FTS is based on complex Langevin dynamics [111, 112], which is related to an earlier formulation for stochastic quantization [113, 114] and has been applied extensively to polymer solutions [110, 115]. Recently, FTS has provided insights into charge-sequence-dependent LLPS of IDRs [81, 87, 88, 107, 116]. The starting point of FTS is identical to that of rG-RPA. FTS invokes no RPA and is thus advantageous over rG-RPA in this regard, though it is still limited by the lattice size used for simulation and its restricted treatment of excluded volume [88]. Here we apply the protocol detailed in refs. [87, 88].

A simple model of ATP-Mg for FTS

We adopt a 6-bead polymeric representation of (ATP-Mg)2- (Fig. 7a) in which four negative and two positive charges serve to model ATP4- and Mg2+ respectively. Modeling (ATP-Mg)2- as a short charged polymer enables application of existing FTS formulations for multiple charge sequences to systems with IDRs and (ATP-Mg)2-. While the model in Fig. 7a does not capture structural details, its charge distribution does correspond roughly to that of the chemical structure of (ATP-Mg)2-. In developing FTS models involving IDR, (ATP-Mg)2- , and NaCl, we first assume for simplicity that (ATP-Mg)2- does not dissociate and consider systems consisting of any given bulk concentrations of IDR and (ATP-Mg)2- wherein all positive and negative charges on the IDR and (ATP-Mg)2- are balanced, respectively, by Cl– and Na+ to maintain overall electric neutrality (Fig. 7a).

FTS rationalizes experimental observation of Caprin1-ATP interactions. (a) The 6-bead model for (ATP-Mg)2- and the single-bead models for monovalent salt ions used in the present FTS. (b-e) Normalized protein-protein correlation functions at three [(ATP-Mg)2-] values (b, c) and protein-ion correlation functions [Eq. (7)] at [(ATP-Mg)2-]/b-3 = 0.03 (d, e) for Caprin1 (b, d) and pY-Caprin1 (c, e), computed for Bjerrum length lB = 7b. Horizontal dashed lines are unity baselines (see text). (f) Values of position-specific integrated correlation Gpq(i)/ρp,i0ρq0 (left vertical axis) correspond to the relative contact frequencies between individual residues (i) along the Caprin1 IDR sequence with q = (ATP-Mg)2- , Na+, or Cl– under the same conditions as (d) [Eq. (S45)] (color symbols). Included for comparison are experimental NMR volume ratios V/V0 data on site-specific Caprin1-ATP association [74]. V/V0 decreases with increased contact probability, although a precise relationship is yet to be determined. Thus, the plotted 1 - V/V0 (grey data points, right vertical scale) is expected to correlate with contact frequency.

Phase behaviors can be probed by FTS density correlation functions

LLPS of FTS systems can be monitored by correlation functions [88]. Here, we compute intra-species IDR self-correlation functions Gpp(r) (Fig. 7b,c) and inter-species crosscorrelation functions Gpq(r) between the IDR and (ATP-Mg)2- or NaCl (Fig. 7d,e) at three different bulk [(ATP-Mg)2-] = 10-4b-3, 0.03b-3, and 0.5b-3, where b may be taken as the peptide virtual bond length ≈ 3.8Å (Materials and Methods). The correlation functions in Fig. 7b-e are normalized by bulk densities ρp0 of the IDR and ρp0 for (ATP-Mg)2-, Na+ or Cl–, wherein density is the bead density for the given molecular species in units of b-3. LLPS of the IDR is signaled by Gpp(r)/(ρp0)2 in Fig. 7b,c dropping below the unity baseline (dashed) at large distance r because it implies a spatial region with depleted IDR below the bulk concentration, which is possible only if the IDR is above the bulk concentration in at least another spatial region. In other words, Gpp(r)/(ρp0)2<1 for large r indicates that IDR concentration is heterogeneous and thus the system is phase separated. For small r, Gpp(r)/(ρp0)2 is generally expected to increase because IDR chain connectivity facilitates correlation among residues local along the chain. On top of this, LLPS propensity may be quantified by Gpp(r)/(ρp0)2 for small r because a higher value indicates a higher tendency for different chains to associate and thus a higher LLPS propensity [88].

FTS rationalizes [ATP-Mg]-modulated Caprin1 reentrant phase behaviors and their colocalization in the condensed phase

[(ATP-Mg)2-]-modulated reentrance is predicted by FTS for Caprin1 but not for pY-Caprin1: When [(ATP-Mg)2-]/b-3 varies from 10-4 to 0.03 to 0.5, small-r values of the Caprin1 Gpp(r) in Fig. 7b initially increase then decrease, whereas the corresponding small-r values of the pY-Caprin1 Gpp(r) in Fig. 7c decrease monotonically, consistent with rG-RPA (Fig. 2g,h,k,l) and experiment (Fig. 3). The inter-species cross-correlations in Fig. 7d,e show further that when an IDR condensed phase is present at [(ATP-Mg)2-] = 0.03b-3 (as indicated by large-r behaviors of Gpp(r)/(ρp0)2 in Fig. 7b,c), (ATP-Mg)2- is colocalized with Caprin1 or pY-Caprin1 (high value of Gpq/ρp0ρq0 for small r) in the IDR-condensed droplet. By comparison, the variation of [Na+] and [Cl–] is much weaker. For Caprin1, Cl– is enhanced over Na+ in the Caprin1 condensed phase (small-r Gpq/ρp0ρq0 of the former larger than the latter in Fig. 7d), but the reverse is seen for pY-Caprin1 (Fig. 7e). This FTS-predicted difference, most likely arising the positive net charge on Caprin1 and the smaller negative net charge on pY-Caprin1, is consistent with the MD results in Fig. 4c,d and SI Appendix, Fig. S3.

FTS rationalizes experimentally observed residue-specific binding of Caprin1 with ATP-Mg

The propensities for (ATP-Mg)2- , Na+ , and Cl– to associate with each residue i along the Caprin1 IDR (i = 1, 2, … , 103) in FTS are quantified by the residue-specific integrated correlation Gpq(i)/ρp,i0ρq0 in Fig. 7f, which is the integral of the corresponding Gpq(i)(r) from r = 0 to a relative short cutoff distance r = rcontact to provide a relative contact frequency for residue i and ionic species q to be in spatial proximity (Materials and Methods and SI Appendix). Notably, the residue-position-dependent integrated correlation for (ATP-Mg)2- varies significantly, exhibiting much larger values near the N-terminal and a little before the C-terminal but weaker correlation elsewhere (Fig. 7f, red symbols). The two regions of high integrated correlation (i.e., favorable association) coincide with regions with high sequence concentration of positively charged residues. This FTS prediction is remarkably similar to the experimental NMR finding that binding between (ATP-Mg)2- and Caprin1 occurs strongly at the arginine-rich N- and C-terminal regions, as indicated by the volume ratio V/V0 data in Fig. 1C of ref. [74] that quantifies the ratio of peaks in NMR spectra in the presence and absence of trace amounts of ATP-Mn. For comparison with the FTS results, this set of experimental data is replotted as 1 - V/V0 in Fig. 7f (grey symbols, right vertical axis) to illustrate the similarity in experimental and theoretical trends because 1 - V/V0 is expected to trend with contact frequency. Corresponding FTS results for Na+ and Cl– in Fig. 7f exhibit much less residue-position-dependent variation, with Cl– displaying only slightly enhanced association in the same arginine-rich regions, and Na+ showing even less variation, presumably because the positive charges on Caprin1 are already essentially neuralized by the locally associated (ATP-Mg)2- or Cl– ions. The theory-experiment agreement in Fig. 7f regarding ATP-Caprin1 interactions indicates once again that electrostatics is an important driving force underlying many aspects of experimentally observed Caprin1-(ATP-Mg)2- association.

FTS snapshots of [ATP-Mg]-modulated reentrant phase behaviors and Caprin1- ATP-Mg colocalization

The above FTS-predicted trends are further illustrated in Fig. 8 by field snapshots. Such FTS snapshots are generally useful for visualization and heuristic understanding [81, 88, 107], including insights into subtler aspects of spatial arrangements exemplified by recent studies of subcompartmentalization entailing either co-mixing or demixing in multiple-component LLPS that are verifable by explicit-chain MD [88, 107]. Now, trends deduced from the correlation functions in Fig. 7 are buttressed by the representative snapshots in Fig. 8: As the bead density of (ATP-Mg)2- is increased from 10-4b-3 to 0.03b-3 to 0.5b-3, the spatial distribution of Caprin1 evolves from an initially dispersed state to a concentrated droplet to a (reentrant) dispersed state again (Fig. 8a), whereas the initial dense pY-Caprin1 droplet becomes increasingly dispersed monotonically (Fig. 8b). Colocalization of (ATP-Mg)2- with both the Caprin1 (Fig. 8c) and pY-Caprin1 (Fig. 8d) droplets is clearly visible at [(ATP-Mg)2-] = 0.03b-3, though the degree of colocalization is appreciably higher for Caprin1 than for pY-Caprin1. This is likely because the positive net charge of Caprin1 is more attractive to (ATP-Mg)2-. By comparison, variations in Na+ and Cl– distribution between Caprin1/pY-Caprin1 dilute and condensed phases are not so discernible in Fig. 8e-h, consistent with the small differences in the corresponding FTS correlation functions (Fig. 7d,e).

FTS rationalizes colocalization of ATP-Mg with the Caprin1 condensate. FTS snapshots are from simulations at lB = 7b (same as that for Fig. 7). Spatial distributions of real positive parts of the density fields for the protein (a, b), (ATP-Mg)2- (c, d), Na+ (e, f), and Cl– (g, h) components are shown by three snapshots each for Caprin1 (left panels) and pY-Caprin1 (right panels) at different [(ATP-Mg)2-] values as indicated. Colocalization of (ATP-Mg)2- with the Caprin1 condensed droplet is clearly seen in the [(ATP-Mg)2-]/b-3 = 0.03 panel of (c).

Robustness of general trends predicted by FTS

We have also assessed the generality of the results in Figs. 7 and 8 by considering three variations in the molecular species treated by FTS: (i) Caprin1 or pY-Caprin1 with only Na+ and Cl– but no (ATP-Mg)2- (SI Appendix, Fig. S4), (ii) Caprin1 with (ATP-Mg)2- and either Na+ or Cl– (but not both) to maintain overall charge neutrality or pY-Caprin1 with (ATP-Mg)2- and Na+ as counterion but no Cl– (SI Appendix, Fig. S5), and (iii) Caprin1 or pY-Caprin1 with ATP4- , Mg2+ , Na+ and Cl– (SI Appendix, Fig. S6). Despite these variations in FTS models, SI Appendix Figs. S4–S6 consistently show reentrant behavior for Caprin1 but not pY-Caprin1 and Figs.S5 and S6 both exhibit colocalization of ATP with condensed Caprin1, suggesting that these features are robust consequences of the basic electrostatics at play in Caprin1/pY-Caprin1 + ATP-Mg + NaCl systems.

Discussion

It is reassuring that, in agreement with experiment, all of our electrostatics-based theoretical approaches consistently predict salt-dependent reentrant phase behaviors for Caprin1, whereas pY-Caprin1 LLPS propensity decreases monotonically with increasing salt (Figs. 2, 4, 7, and 8). This effect applies to small monovalent salts exemplified by Na+ and Cl– as well as to our electrostatics-based models of (ATP-Mg)2- or ATP4- , with ATP exhibiting a significant colocalization with the Caprin1 condensed phase (Figs. 2g,h,k,l and 8c).

Related studies of electrostatic effects on biomolecular condensates

Our theoretical predictions are also largely in agreement with recent computational studies on salt concentrations in the dilute versus condensed phases [108] and salt-dependent reentrant behaviors [42] of other biomolecular condensates, including explicit-water, explicit-ion atomic simulations with preformed condensates of the N-terminal RGG domain of LAF-1 [108] and of the highly positive proline-arginine 25-repeat dipeptide PR25 [117].

A recent study examines salt-dependent reentrant LLPSs of full-length FUS (WT and G156E mutant), TDP-43, bromodomain-containing protein 4 (Brd4), sex-determining region Y-box 2 (Sox2), and annexin A11 [42]. Unlike the requirement of a nonzero monovalent salt concentration for Caprin1 LLPS, LLPS is observed for all these six proteins with KCl, NaCl or other salts at concentrations as low as 50 mM. Also unlike Caprin1, their protein condensates dissolve at intermediate salt then re-appear at higher salt, a phenomenon the authors rationalize by a tradeoff between decreasing favorability of cation-anion interactions and increasing favorability of cation-cation, cation-π, hydrophobic, and other interactions with increasing monovalent salt [42].

Two reasons may account for this difference. First, Caprin1 does not phase separate at low salt because it is a relatively strong polyelectrolyte (NCPR = +13/103 = +0.126). By comparison, five of the six proteins in ref. [42] are much weaker polyelectrolytes or not at all, with NCPR = +14/526 = +0.0266, +13/526 = +0.0247, -7/80 = -0.0875, 0, and +3/326 = +0.00920, respectively, for FUS (WT, mutant), TDP-43, Brd4, and A11. Apparently, their weak electrostatic repulsions can be overcome by favorable nonelectrostatic interactions alone to enable LLPS.

Second, compared to Caprin1, the proteins in ref. [42] are either significantly larger (WT and mutant FUS) or significantly more hydrophobic and aromatic (the other four proteins), both properties are conducive to LLPS. For instance, although Sox2’s NCPR = +14/88 = +0.159 is higher than that of Caprin1, among Sox2’s amino acid residues, 21/88 = 23.9% are large hydrophobic or aromatic residues leucine (L), isoleucine (I), valine (V), methionine (M), phenylalanine (F), or tryptophan (W), and 17/88 = 19.3% are large aliphatic residues L, I, V, or M. This amino acid composition suggests that hydrophobic or n-related interactions in Sox2 can be sufficient to overcome electrostatic repulsion to effectuate LLPS at zero salt. In contrast, the Caprin1 IDR contains merely one L; only 10/103 = 9.7% of the residues of Caprin1 are in the L, I, V, M, F, W hydrophobic/aromatic category and only 6/103 = 5.8% are in the L, I, V, M aliphatic category. The corresponding aliphatic fractions of TDP-43, Brd4 and A11, at 21/80 = 26.3%, 33/132 = 25%, and 90/326 = 27.6%, respectively, are also significantly higher than that of Caprin1.

Effects of salt on biomolecular LLPS

Effects of salts on LLPS, including partition of salt into polymer-rich phases, are of long-standing interest in polymer physics [118]. In the biomolecular condensate context, the versatile functional roles of salts are highlighted by the interplay between electrostatic and cation-π interactions [119, 120], salts’ modulating effects on heat-induced LLPSs of RNAs [121], their regulation of condensate liquidity [122], and even their potential impact in extremely high-salt exobiological environments [123]. While some of these recent studies focus primarily on salts’ electrostatic screening effects without changing the signs of the effective polymer charge-charge interaction [119], effective attractions between like charges bridged by salt or other oppositely-charged ions [94] as illustrated by Caprin1 (Fig. 5f) are likely needed to account for phenomena such as salt-induced dimerization of highly charged, medically relevant arginine-rich cell-penetrating short peptides [124, 125].

Tielines in protein-salt phase diagrams

In view of Caprin1’s polyelectrolytic nature, the mildly negative tieline slopes in Fig. 2a,b are consistent with rG-RPA predictions for a fully charged polyelectrolyte (Fig. 10a of ref. [62]). This depletion of monovalent salt in the condensed phase is similar to that observed in the complex coacervation of oppositely charged polyelectrolytes [126–128]. By comparison, the positive rG-RPA tieline slopes for polyampholytic pY-Caprin1 (Fig. 2c,d), confirmed by MD in Fig. 4d, are appreciably steeper than that predicted for fully charged (±1) diblock polyampholytes by rG-RPA and the essentially flat tielines predicted by FTS (Fig. 10b of ref. [62] and Fig. 7 of ref. [82]). Whether this difference originates from the presence of divalently charged (-2) phosphorylated sites in pY-Caprin1 remains to be elucidated. In any event, tieline analysis is generally instrumental for revealing details, such as stoichiometry, of the interactions driving multiple-component biomolecular LLPSs [14, 129], rG-RPA should be broadly useful as a computationally efficient tool for this purpose [62].

Counterion valency

Our rG-RPA prediction that the maximum condensed-phase [Caprin1] at low [Na+] is substantially higher with divalent than with monovalent counterions is in line with early findings that higher-valency counterions are more effective in bridging polyelectrolyte interactions to favor LLPS [130] and recent observations that salt ions with higher valencies enhance biomolecular LLPS [131, 132]. The possibility that this counterion/salt effect on LLPS may be exploited more generally for biological functions and/or biomedical applications remains to be further explored. In this regard, while recognizing that ATP can engage in n-related interactions [56–58], our electrostatics-based perspective of ATP-dependent reentrant phase behaviors is consistent with recent observations on polylysine LLPS modulated by enzymatically catalyzed ATP turnovers [128, 133].

Prospective extensions of the present theoretical methodology

Beyond the above comparisons, further experimental testing of other aspects of our theoretical predictions should be pursued, especially those pertaining to pY-Caprin1. Future theoretical efforts should address a broader range of scenarios by independent variations of [ATP4-], [Mg2+], [Na+], [Cl–] and to account for nonelectrostatic aspects of ATP-Mg dissociation [134] with predictions such as tieline slopes analyzed in detail to delineate effects of configurational entropy of salt ions [135] and solvent quality [136]. In addition to our basic modeling constructs, the impact of excluded volume and solvent/cosolute-mediated temperature-dependent effective interactions should be incorporated. Excluded volume is known to affect LLPS [82], demixing of IDP species in condensates [88], and partition of salt ions in polymer LLPS [127]. Moreover, LCST can be driven not only by hydrophobicity [4, 10, 21] but also by electrostatics, as suggested by experiment on complex coacervates of oppositely charged polyelectrolytes [137]. Bringing together these features into a comprehensive formulation will afford a more accurate physical picture.

Summary

To recapitulate, we have employed three complementary theoretical and computational approaches to account for the interplay between sequence pattern, phosphorylation, counterion, and salt in the phase behaviors of IDPs. Application to the Caprin1 IDR and its phosphorylated variant pY-Caprin1 provides physical rationalization for a variety of trends observed in experiments, including reentrance behaviors and very substantial ATP colocalization. These findings support a significant—albeit not exclusive—role of electrostatics in these biophysical phenomena, providing physical insights into effects of sequence-specific charge-charge interactions on ATP-modulated physiological functions of biomolecular condensates such as regulation of ion concentrations. The approach developed here should be of general utility as a computationally efficient tool for hypothesis generation, design of new experiments, exploration and testing of biophysical scenarios, as well as a starting point for more sophisticated theoretical/computational modeling.

Materials and Methods

Further details of the experimental and theoretical/computational methodologies outlined below are provided in SI Appendix.

Experimental sample preparation

The low complexity 607–709 domain of Caprin1 was expressed and purified as before [28, 74]. WT Caprin1 was used in all experiments except those on [NaCl] dependence reported in Table 1 and Fig. 3a, for which a double mutant was used because residue pairs N623-G624 and N630-G631 in WT Caprin1 form isoaspartate (IsoAsp) glycine linkages over time which alters the charge distribution of the IDR [73].

Phosphorylation of the Caprin1 IDR

Phosphorylation of the WT Caprin1 IDR was performed as described in our prior study [28] by using the kinase domain of mouse Eph4A (587–896) [138] with an N-terminal His-SUMO tag.

Determination of phase diagrams

We established phase diagrams for Caprin1 and pY-Caprin1 by measuring the protein concentrations in dilute and condensed phases across a range of [NaCl]s (Fig. 1c,d). Initially homogenizing the two phases of the demixed samples into a milky dispersion through vortexing, ~ 200 μL aliquots were then incubated in a PCR thermocycler with a heated lid at 90°C, in triplicate, for a minimum of one hour. During incubation, the condensed phase settled and formed a clear phase at the bottom. For concentration measurements, the samples were diluted in 6 M GdmCl and 20 mM NaPi (pH 6.5). The dilute phase (top layer) was analyzed through a tenfold dilution of 10 μL samples, and the condensed phase (bottom layer) was analyzed through 250- to 500-fold dilution of 2 or 10 μL samples.

Concentrations of salt and ATP-Mg in dilute and condensed phases

Inductively coupled plasma optical emission spectroscopy (ICP-OES) measurements of [Na+] were performed using a Thermo Scientific iCAP Pro ICP-OES instrument in axial mode. ICP-OES was also used to determine [ATP] and [Mg2+] (Table 2). The detection of phosphorus and magnesium served as proxies for quantifying ATP and Mg2+ levels, respectively. Standard curves were prepared using solutions with known [ATP] and [Mg2+], ranging from 0 to 90 ppm for ATP and 0 to 25 ppm for Mg2+.

Caprin1 phase separation propensity at high salt concentrations

A 6 mM solution of double-mutant Caprin1 IDR (see above) in buffer (25 mM sodium phosphate, pH 7.4) was prepared by exchanging (3 times) the purified protein after size exclusion chromatography using centrifugal concentrators (3 kDa, EMD Millipore). Caprin samples for turbidity measurements were prepared by taking 0.5 μL of the above solution and diluting it into buffer (25 mM sodium phosphate, pH 7.4) containing varying [NaCl]s ranging from 0 to 4.63 M, in a sample volume of 9 μL, so as to achieve [Caprin1] of 300 μM. After rigorous mixing, 5 μL samples were loaded into a μCuvette G1.0 (Eppendorf). OD600 measurements (Fig. 3a) were recorded three times using a BioPhotometer D30 (Eppendorf).

[ATP-Mg]-dependent Caprin1 phase behaviors

Turbidity assays were conducted using the method we described previously [73].

Sequence-specific theory of heteropolymer phase separation

As detailed in refs. [87, 107], an example of the sequence-specific polymer theories [16, 62] is that for a solution with a single species of charged heteropolymers in np copies, nc counterions (same type), and ns salt ions (same type, but different from the counterions). Each polymer chain has N monomers (residues) with charge sequence |σ〉=[ σ1,σ2…σN ]T in vector notation, where σi ∈ {0, ±1, -2} is the charge of the ith residue. The counterions and salt ions are monomers carrying zc and zs charges, respectively. The particle-based partition function is given by Ζ=1nP!nc!ns!nw!∫∏α=1nP∏i=1NdRα,i∏α=1ns+ncdrae−T[ R ]−U[ R,r ], (2)

where nw denotes the number of water (solvent) molecules, Rα,i is the position vector of the ith residue of the ath polymer, ra is the position vector of the ath small ion. T accounts for polymer chain connectivity modeled by a Gaussian elasticity potential with Kuhn length l. U describes the interactions among all molecular components of the system, here consisting only of Coulomb electrostatics (el) and excluded-volume (ex) for simplicity, viz., U = Uel + Uex. Their interaction strengths are governed by the Bjerrum length lB and the two-body excluded volume parameter v2. By introducing conjugate fields ψ(r), w(r) and applying the Hubbard-Stratonovich transformation, the system defined by the particle-based partition function in Eq. (2) is recast as a field theory of ψ, w in which their interactions with polymer, salt, and counterion are described, respectively, by single molecule partition functions Qp, Qs, and Qc. For instance, Qp[ ψ,w ]=∫∏i=1NdRiexp(−HP[ R;ψ,w ]) (3)

where Hp is the single-polymer Hamiltonian and the chain label α is dropped.

Renormalized-Gaussian random-phase-approximation (rG-RPA)

Following refs. [37, 62], Hp can be separated into a Gaussian-chain Hamiltonian with an effective (renormalized) Kuhn length l1 = xl and a remaining term, Hp=Hp0+Hp1 , where HP0=32l2x∑i=1N−1(Ri+1−Ri)2, (4a) HP1=32l2(1−1x)∑i=1N−1(Ri+1−Ri)2−1∑i=1N[ σiψ(Ri)+w(Ri) ], (4b)

with i2 = -1. By requiring the observable polymer square end-to-end distance be properly quantified by Hp0 , x can be approximated by variational theory [37]. RPA can then be applied to the renormalized Gaussian (rG) chain system with l → xl and a corresponding scaling of the contour length to arrive at an improved theory, rG-RPA, for sequence-specific LLPS.

Explicit-ion coarse-grained molecular dynamics (MD)

The MD model in this work augments a class of implicit-water coarse-grained models [78, 84] that utilize a “slab” approach for efficient equilibration [102] by incorporating explicit small ions. As before [84], the total MD potential energy UT is the sum of long-spatial-range electrostatic (el) and short-spatial-range (sr) interactions of the Lennard-Jones (LJ) type as well as bond interactions, i.e., UT = Uel + Usr + Ubond. With small ions, the electrostatic component is given by a sum of polymer-polymer (pp), polymer-ion (pi), and ion-ion (ii) contributions: Uel = Uel,pp + Uel,pi + Uel,ii. Details of these terms are provided in SI Appendix.

Field-theoretic simulation (FTS)

FTS is useful for sequence-specific multiple-component LLPSs encountered in biomolecular settings. The new applications developed here are based on recent advances (see, e.g., refs. [81, 82, 87, 88, 110, 115]). Consider the field theoretic Hamiltonian H[ w,ψ ]=∫dr([ ∇ψ(r) ]28πlB+w(r)22v2)−∑mnmlnQm[ w⌣,ψ⌣ ], (5)

where Qm is single-molecule partition function [here m labels the components in the system, cf. Eq. (3)] and the breves denote convolution with Γ, i.e., for a generic field ϕ, ϕ⌣(r)= Γ*ϕ(r)≡∫dr′Γ(r−r′)ϕ(r′) ; here ϕ = w, ψ, and Γ is a Gaussian smearing function [87]. FTS utilizes the Complex-Langevin (CL) method [111, 112] by introducing an artifical CL time variable (t), viz., w(r) → w(r,t), ψ(r) → ψ(r,t) and letting the system evolve in CL time in accordance with a collection of Langevin equations ∂ϕ(r,t)∂t=−δHδϕ(r,t)+ηϕ(r,t), ϕ=w,ψ, (6)

where the Gaussian noise ηϕ(r, t) satisfies 〈 ηϕ(r,t)ηϕ′(r′,t′) 〉=2δϕ,ϕ′δ(r−r′)δ(t−t′) . Thermal averages of thermodynamic observables are then computed as asymptotic CL time averages of the corresponding field operators. Spatial information about condensation and proximity of various components is readily gleaned from density-density correlation functions [87, 88], Gm,n(| r−r′ |)=〈 ρ^m(r)ρ^n(r′) 〉, (7)

where m, n are labels for the components in the model system. For instance, m may represent all polymer beads (denoted “p”) irrespective of the sequence positions of the beads [ ρ^p(r)=∑i=1np∑α=1NΓ(r−Rα,i) ] , and n may represent all six beads in our ATP-Mg model (Fig. 7a). One may also define Gpq(i)(| r−r′ |)≡〈 ρ^P,i(r)ρ^q(r′) 〉, (8)

where (i) represents the ith residue along a protein chain [ ρ^p,i(r)≡∑α=1NΓ(r−Rα,i) ] , and q = (ATP-Mg)2- , Na+, or Cl–. With this definition, residue-specific relative contact frequencies are estimated by integrating Eq. (8) over a spherical volume within a small inter-component distance rcontact: Gpq(i)≡4π∫0rcontactdr r2Gpq(i)(r). (9)

For the normalized Gpq(i)/ρp,i0ρq0 plotted in Fig. 7f, ρp,i0 and ρq0 are bulk (overall) densities, respectively, of the ith protein residue and of (ATP-Mg)2- or small ions, and rcontact ≈ 1.5b is used to characterize contacts. Further details are provided in SI Appendix.

eLife Assessment

In this important study, the authors employed three types of theoretical/computational models (coarse-grained molecular dynamics, analytical theory and field-theoretical simulations) to analyze the impact of salt on protein liquid-liquid phase separation. These different models reinforce each other and together provide convincing evidence to explain distinct salt effects on ATP mediated phase separation of different variants of caprin1. The insights and general approach are broadly applicable to the analysis of protein phase separation.

Reviewer #1 (Public Review):

Summary:

The authors used multiple approaches to study salt effects in liquid-liquid phase separation (LLPS). Results on both wild-type Caprin1 and mutants and on different types of salts contribute to a comprehensive understanding.

Strengths:

The main strength of this work is the thoroughness of investigation. This aspect is highlighted by the multiple approaches used in the study, and reinforced by the multiple protein variants and different salts studied.

Weaknesses:

(1) The multiple computational approaches are a strength, but they're cruder than explicit-solvent all-atom molecular dynamics (MD) simulations and may miss subtle effects of salts. In particular, all-atom MD simulations demonstrate that high salt strengthens pi-types of interactions (ref. 42 and MacAinsh et al, https://www.biorxiv.org/content/10.1101/2024.05.26.596000v3).

(2) The paper can be improved by distilling the various results into a simple set of conclusions. By example, based on salt effects revealed by all-atom MD simulations, MacAinsh et al. presented a sequence-based predictor for classes of salt dependence. Wild-type Caprin1 fits right into the "high net charge" class, with a high net charge and a high aromatic content, showing no LLPS at 0 NaCl and an increasing tendency of LLPS with increasing NaCl. In contrast, pY-Caprin1 belongs to the "screening" class, with a high level of charged residues and showing a decreasing tendency of LLPS.

(3) Mechanistic interpretations can be further simplified or clarified. (i) Reentrant salt effects (e.g., Fig. 4a) are reported but no simple explanation seems to have been provided. Fig. 4a,b look very similar to what has been reported as strong-attraction promotor and weak-attraction suppressor, respectively (ref. 50; see also PMC5928213 Fig. 2d,b). According to the latter two studies, the "reentrant" behavior of a strong-attraction promotor, CL- in the present case, is due to Cl-mediated attraction at low to medium [NaCl] and repulsion between Cl- ions at high salt. Do the authors agree with this explanation? If not, could they provide another simple physical explanation? (ii) The authors attributed the promotional effect of Cl- to counterion-bridged interchain contacts, based on a single instance. There is another simple explanation, i.e., neutralization of the net charge on Caprin1. The authors should analyze their simulation results to distinguish net charge neutralization and interchain bridging; see MacAinsh et al.

(4) The authors presented ATP-Mg both as a single ion and as two separate ions; there is no explanation of which of the two versions reflects reality. When presenting ATP-Mg as a single ion, it's as though it forms a salt with Na+. I assume NaCl, ATP, and MgCl2 were used in the experiment. Why is Cl- not considered? Related to this point, it looks like ATP is just another salt ion studied and much of the Results section is on NaCl, so the emphasis of ATP ("Diverse Roles of ATP" in the title is somewhat misleading.

Reviewer #2 (Public Review):

Summary:

In this paper, Lin and colleagues aim to understand the role of different salts on the phase behavior of a model protein of significant biological interest, Caprin1, and its phosphorylated variant, pY-Caprin1. To achieve this, the authors employed a variety of methods to complement experimental studies and obtain a molecular-level understanding of ion partitioning inside biomolecular condensates. A simple theory based on rG-RPA is shown to capture the different salt dependencies of Caprin1 and pY-Caprin1 phase separation, demonstrating excellent agreement with experimental results. The application of this theory to multivalent ions reveals many interesting features with the help of multicomponent phase diagrams. Additionally, the use of CG model-based MD simulations and FTS provides further clarity on how counterions can stabilize condensed phases.

Strengths:

The greatest strength of this study lies in the integration of various methods to obtain complementary information on thermodynamic phase diagrams and the molecular details of the phase separation process. The authors have also extended their previously proposed theoretical approaches, which should be of significant interest to other researchers. Some of the findings reported in this paper, such as bridging interactions, are likely to inspire new studies using higher-resolution atomistic MD simulations.

Weaknesses:

The paper does not have any major issues.

Reviewer #3 (Public Review):

Authors first use rG-RPA to reproduce two observed trends. Caprin1 does not phase separate at very low salt but then undergoes LLPS with added salt while further addition of salt reduces its propensity to LLPS. On the other hand pY-Caprin1 exhibits a monotonic trend where the propensity to phase separate decreases with the addition of salt. This distinction is captured by a two component model and also when salt ions are explicitly modeled as a separate species with a ternary phase diagram. The predicted ternary diagrams (when co and counter ions are explicitly accounted for) also predict the tendency of ions to co-condense or exclude proteins in the dense phase. Predicted trends are generally in line with the measurement for Cparin1. Next, the authors seek to explain the observed difference in phase separation when Arginines are replaced by Lysines creating different variants. In the current rG-RPA type models both Arginine (R) and Lysine (K) are treated equally since non-electrostatic effects are only modeled in a mean-field manner that can be fitted but not predicted. For this reason, coarse grain MD simulation is suitable. Moreover, MD simulation affords structural features of the condensates. They used a force field that is capable of discriminating R and K. The MD predicted degrees of LLPS of these variants again is consistent with the measurement. One additional insight emerges from MD simulations that a negative ion can form a bridge between two positively charged residues on the chain. These insights are not possible to derive from rG-RPA. Both rG-RPA and MD simulation become cumbersome when considering multiple types of ions such as Na, Cl, [ATP] and [ATP-Mg] all present at the same time. FTS is well suited to handle this complexity. FTS also provides insights into the co-localization of ions and proteins that is consistent with NMR. By using different combinations of ions they confirm the robustness of the prediction that Caprin1 shows salt-dependent reentrant behavior, adding further support that the differential behavior of Caprin1, and pY-Caprin1 is likely to be mediated by charge-charge interactions.

Author response:

The following is the authors’ response to the current reviews.

Reviewer #1 (Public Review):

Comments on revisions:

This revision addressed all my previous comments.

Reviewer #3 (Public Review):

Comments on revisions:

The authors addressed my comments and it is ready for publication.

We are grateful for the reviewers’ effort and are encouraged by their generally positive assessment of our manuscript.

Reviewer #1 (Recommendations For The Authors):

This revision addressed all my previous comments. The only new issue concerns the authors’ response to the following comment of reviewer 3:

(2) Authors note ”monovalent positive salt ions such as Na+ can be attracted, somewhat counterintuitively, into biomolecular condensates scaffolded by positively-charged polyelectrolytic IDRs in the presence of divalent counterions”. This may be due to the fact that the divalent negative counterions present in the dense phase (as seen in the ternary phase diagrams) also recruit a small amount of Na+.

Author reply: The reviewer’s comment is valid, as a physical explanation for this prediction is called for. Accordingly, the following sentence is added to p. 10, lines 27-29: ...

Here are my comments on this issue. Most IDPs with a net positive charge still have negatively charged residues, which in theory can bind cations. In fact, Caprin1 has 3 negatively charged residues (same as A1-LCD). All-atom simulations of MacAinsh et al (ref 72) have shown that these negatively charged residues bind Na+; I assume this effect can be captured by the coarsegrained models in the present study. Moreover, all-atom simulations showed that Na+ has a strong tendency to be coordinated by backbone carbonyls, which of course are present on all residues. Suggestions:

(a) The authors may want to analyze the binding partners of Na+. Are they predominantly the3 negatively charged residues, or divalent counterions, or both?

(b) The authors may want to discuss the potential underestimation of Na+ inside Caprin1 condensates due to the lack of explicit backbone carbonyls that can coordinate Na+ in their models. A similar problem applies to backbone amides that can coordinate anions, but to a lesser extent (see Fig. 3A of ref 72).

The reviewer’s comments are well taken. Regarding the statement in the revised manuscript “This phenomenon arises because the positively charge monovalent salt ions are attracted to the negatively charged divalent counterions in the protein-condensed phase.”, it should be first noted that the statement was inferred from the model observation that Na+ is depleted in condensed Caprin1 (Fig. 2a) when the counterion is monovalent (an observation that was stated almost immediately preceding the quoted statement). To make this logical connection clearer as well as to address the reviewer’s point about the presence of negatively charged residues in Caprin1, we have modified this statement in the Version of Record (VOR) as follows:

“This phenomenon most likely arises from the attraction of the positively charge monovalent salt ions to the negatively charged divalent counterions in the proteincondensed phase because although the three negatively charged D residues in Caprin1 can attract Na+, it is notable that Na+ is depleted in condensed Caprin1 when the counterion is monovalent (Fig. 2a).”

The reviewer’s suggestion (a) of collecting statistics of Na+ interactions in the Caprin1 condensate is valuable and should be attempted in future studies since it is beyond the scope of the present work. Thus far, our coarse-grained molecular dynamics has considered only monovalent Cl− counterions. We do not have simulation data for divalent counterions.

Following the reviewer’s suggestion (b), we have now added the following sentence in Discussion under the subheading “Effects of salt on biomolecular LLPS”:

“In this regard, it should be noted that positively and negatively charged salt ions can also coordinate with backbone carbonyls and amides, respectively, in addition to coordinating with charged amino acid sidechains (MacAinsh et al., eLife 2024). The impact of such effects, which are not considered in the present coarse-grained models, should be ascertained by further investigations using atomic simulations (MacAinsh et al., eLife 2024; Rauscher & Pom`es, eLife 2017; Zheng et al., J Phys Chem B 2020).”

Here we have added a reference to Rauscher & Pom`es, eLife 2017 to more accurately reflect progress made in atomic simulations of biomolecular condensates.

More generally, regarding the reviewer’s comments on the merits of coarse-grained versus atomic approaches, we re-emphasize, as stated in our paper, that these approaches are complementary. Atomic approaches undoubtedly afford structurally and energetically high-resolution information. However, as it stands, simulations of the assembly-disassembly process of biomolecular condensate are nonideal because of difficulties in achieving equilibration even for a small model system with < 10 protein chains (MacAinsh et al., eLife 2024) although well-equilibrated simulations are possible for a reasonably-sized system with ∼ 30 chains when the main focus is on the condensed phase (Rauscher & Pom`es, eLife 2017). In this context, coarse-grained models are valuable for assessing the energetic role of salt ions in the thermodynamic stability of biomolecular condensates of physically reasonable sizes under equilibrium conditions.

In addition to the above minor additions, we have also added citations in the VOR to two highly relevant recent papers: Posey et al., J Am Chem Soc 2024 for salt-dependent biomolecular condensation (mentioned in Dicussion under subheadings “Tielines in protein-salt phase diagrams” and “Counterion valency” together with added references to Hribar et al., J Am Chem Soc 2002 and Nostro & Ninham, Chem Rev 2012 for the Hofmeister phenomena discussed by Posey et al.) and Zhu et al., J Mol Cell Biol 2024 for ATP-modulated reentrant behavior (mentioned in Introduction). We have also added back a reference to our previous work Lin et al., J Mol Liq 2017 to provide more background information for our formulation.

Reviewer #2 (Recommendations For The Authors):

The authors have done a great job addressing previous comments.

We thank this reviewer for his/her effort and are encouraged by the positive assessment of our revised manuscript.

---

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

The authors used multiple approaches to study salt effects in liquid-liquid phase separation (LLPS). Results on both wild-type Caprin1 and mutants and on different types of salts contribute to a comprehensive understanding.

Strengths:

The main strength of this work is the thoroughness of investigation. This aspect is highlighted by the multiple approaches used in the study, and reinforced by the multiple protein variants and different salts studied.

We are encouraged by this positive overall assessment.

Weaknesses: (1) The multiple computational approaches are a strength, but they’re cruder than explicit-solvent all-atom molecular dynamics (MD) simulations and may miss subtle effects of salts. In particular, all-atom MD simulations demonstrate that high salt strengthens pi-types of interactions (ref. 42 and MacAinsh et al, https://www.biorxiv.org/content/10.1101/2024.05.26.596000v3).

The relative strengths and limitations of coarse-grained vs all-atom simulation are now more prominently discussed beginning at the bottom of p. 5 through the first 8 lines of p. 6 of the revised manuscript (page numbers throughout this letter refer to those in the submitted pdf file of the revised manuscript), with MacAinsh et al. included in this added discussion (cited as ref. 72 in the revised manuscript). The fact that coarse-grained simulation may not provide insights into more subtle structural and energetic effects afforded by all-atom simulations with regard to π-related interaction is now further emphasized on p. 11 (lines 23–30), with reference to MacAinsh et al. as well as original ref. 42 (Krainer et al., now ref. 50 in the revised manuscript).

(2) The paper can be improved by distilling the various results into a simple set of conclusions. By example, based on salt effects revealed by all-atom MD simulations, MacAinsh et al. presented a sequence-based predictor for classes of salt dependence. Wild-type Caprin1 fits right into the “high net charg”e class, with a high net charge and a high aromatic content, showing no LLPS at 0 NaCl and an increasing tendency of LLPS with increasing NaCl. In contrast, pY-Caprin1 belongs to the “screening” class, with a high level of charged residues and showing a decreasing tendency of LLPS.

This is a helpful suggestion. We have now added a subsection with heading “Overview of key observations from complementary approaches” at the beginning of the “Results” section on p. 6 (lines 18–37) and the first line of p. 7. In the same vein, a few concise sentences to summarize our key results are added to the first paragraph of “Discussion” (p. 18, lines 23– 26). In particular, the relationship of Caprin1 and pY-Caprin1 with the recent classification by MacAinsh et al. (ref. 72) in terms of “high net charge” and “screening” classes is now also stated, as suggested by this reviewer, on p. 18 under “Discussion” (lines 26–30).

(3) Mechanistic interpretations can be further simplified or clarified. (i) Reentrant salt effects (e.g., Fig. 4a) are reported but no simple explanation seems to have been provided. Fig. 4a,b look very similar to what has been reported as strong-attraction promotor and weak-attraction suppressor, respectively (ref. 50; see also PMC5928213 Fig. 2d,b). According to the latter two studies, the “reentrant” behavior of a strong-attraction promotor, CL- in the present case, is due to Cl-mediated attraction at low to medium [NaCl] and repulsion between Cl- ions at high salt. Do the authors agree with this explanation? If not, could they provide another simple physical explanation? (ii) The authors attributed the promotional effect of Cl- to counterionbridged interchain contacts, based on a single instance. There is another simple explanation, i.e., neutralization of the net charge on Caprin1. The authors should analyze their simulation results to distinguish net charge neutralization and interchain bridging; see MacAinsh et al.

The relationship of Cl− in bridging and neutralizing configurations, respectively, with the classification of “strong-attraction promoter” and “weak-attraction suppressor” by Zhou and coworkers is now stated on p. 13 (lines 29–31), with reference to original ref. 50 by Ghosh, Mazarakos & Zhou (now ref. 59 in the revised manuscript) as well as the earlier patchy particle model study PMC5928213 by Nguemaha & Zhou, now cited as ref. 58 in the revised manuscript. After receiving this referee report, we have conducted an extensive survey of our coarse-grained MD data to provide a quantitative description of the prevalence of counterion (Cl−) bridging interactions linking positively charged arginines (Arg+s) on different Caprin1 chains in the condensed phase (using the [Na+] = 0 case as an example). The newly compiled data is reported under a new subsection heading “Explicit-ion MD offers insights into counterion-mediated interchain bridging interactions among condensed Caprin1 molecules” on p. 12 (last five lines)–p. 14 (first 10 lines) [∼ 1_._5 additional page] as well as a new Fig. 6 to depict the statistics of various Arg+–Cl−–Arg+ configurations, with the conclusion that a vast majority (at least 87%) of Cl− counterions in the Caprin1-condensed phase engage in favorable condensation-driving interchain bridging interactions.

(4) The authors presented ATP-Mg both as a single ion and as two separate ions; there is no explanation of which of the two versions reflects reality. When presenting ATP-Mg as a single ion, it’s as though it forms a salt with Na+. I assume NaCl, ATP, and MgCl2 were used in the experiment. Why is Cl- not considered? Related to this point, it looks like ATP is just another salt ion studied and much of the Results section is on NaCl, so the emphasis of ATP (“Diverse Roles of ATP” in the title is somewhat misleading.

We model ATP and ATP-Mg both as single-bead ions (in rG-RPA) and also as structurally more realistic short multiple-bead polymers (in field-theoretic simulation, FTS). We have now added discussions to clarify our modeling rationale in using and comparing different models for ATP and ATP-Mg, as follows:

p. 8 (lines 19–36):

“The complementary nature of our multiple methodologies allows us to focus sharply on the electrostatic aspects of hydrolysis-independent role of ATP in biomolecular condensation by comparing ATP’s effects with those of simple salt. Here, Caprin1 and pY-Caprin1 are modeled minimally as heteropolymers of charged and neutral beads in rG-RPA and FTS. ATP and ATP-Mg are modeled as simple salts (singlebead ions) in rG-RPA whereas they are modeled with more structural complexity as short charged polymers (multiple-bead chains) in FTS, though the latter models are still highly coarse-grained. Despite this modeling difference, rG-RPA and FTS both rationalize experimentally observed ATP- and NaCl-modulated reentrant LLPS of Caprin1 and a lack of a similar reentrance for pY-Caprin1 as well as a prominent colocalization of ATP with the Caprin1 condensate. Consistently, the same contrasting trends in the effect of NaCl on Caprin1 and pY-Caprin1 are also seen in our coarse-grained MD simulations, though polymer field theories tend to overestimate LLPS propensity [99]. The robustness of the theoretical trends across different modeling platforms underscores electrostatics as a significant component in the diverse roles of ATP in the context of its well-documented ability to modulate biomolecular LLPS via hydrophobic and π-related effects [63, 65, 67].”

Here, the last sentence quoted above addresses this reviewer’s question about our intended meaning in referring to “diverse roles of ATP” in the title of our paper. To make this point even clearer, we have also added the following sentence to the Abstract (p. 2, lines 12–13):

“... The electrostatic nature of these features complements ATP’s involvement in π-related interactions and as an amphiphilic hydrotrope, ...”

Moreover, to enhance readability, we have now added pointers in the rG-RPA part of our paper to anticipate the structurally more complex ATP and ATP-Mg models to be introduced subsequently in the FTS part, as follows:

p. 9 (lines 13–15):

“As mentioned above, in the present rG-RPA formulation, (ATP-Mg)2− and ATP4− are modeled minimally as a single-bead ion. They are represented by charged polymer models with more structural complexity in the FTS models below.”

p. 11 (lines 8–11):

These observations from analytical theory will be corroborated by FTS below with the introduction of structurally more realistic models of (ATP-Mg) 2−, ATP4− together with the possibility of simultaneous inclusion of Na+, Cl−, and Mg2+ in the FTS models of Caprin1/pY-Caprin1 LLPS systems.

Reviewer #2 (Public Review):

Summary:

In this paper, Lin and colleagues aim to understand the role of different salts on the phase behavior of a model protein of significant biological interest, Caprin1, and its phosphorylated variant, pY-Caprin1. To achieve this, the authors employed a variety of methods to complement experimental studies and obtain a molecular-level understanding of ion partitioning inside biomolecular condensates. A simple theory based on rG-RPA is shown to capture the different salt dependencies of Caprin1 and pY-Caprin1 phase separation, demonstrating excellent agreement with experimental results. The application of this theory to multivalent ions reveals many interesting features with the help of multicomponent phase diagrams. Additionally, the use of CG model-based MD simulations and FTS provides further clarity on how counterions can stabilize condensed phases.

Strengths:

The greatest strength of this study lies in the integration of various methods to obtain complementary information on thermodynamic phase diagrams and the molecular details of the phase separation process. The authors have also extended their previously proposed theoretical approaches, which should be of significant interest to other researchers. Some of the findings reported in this paper, such as bridging interactions, are likely to inspire new studies using higher-resolution atomistic MD simulations.

Weaknesses:

The paper does not have any major issues.

We are very encouraged by this reviewer’s positive assessment of our work.

Reviewer #3 (Public Review):

Authors first use rG-RPA to reproduce two observed trends. Caprin1 does not phase separate at very low salt but then undergoes LLPS with added salt while further addition of salt reduces its propensity to LLPS. On the other hand pY-Caprin1 exhibits a monotonic trend where the propensity to phase separate decreases with the addition of salt. This distinction is captured by a two component model and also when salt ions are explicitly modeled as a separate species with a ternary phase diagram. The predicted ternary diagrams (when co and counter ions are explicitly accounted for) also predict the tendency of ions to co-condense or exclude proteins in the dense phase. Predicted trends are generally in line with the measurement for Cparin1 [sic]. Next, the authors seek to explain the observed difference in phase separation when Arginines are replaced by Lysines creating different variants. In the current rG-RPA type models both Arginine (R) and Lysine (K) are treated equally since non-electrostatic effects are only modeled in a meanfield manner that can be fitted but not predicted. For this reason, coarse grain MD simulation is suitable. Moreover, MD simulation affords structural features of the condensates. They used a force field that is capable of discriminating R and K. The MD predicted degrees of LLPS of these variants again is consistent with the measurement. One additional insight emerges from MD simulations that a negative ion can form a bridge between two positively charged residues on the chain. These insights are not possible to derive from rG-RPA. Both rG-RPA and MD simulation become cumbersome when considering multiple types of ions such as Na, Cl, [ATP] and [ATP-Mg] all present at the same time. FTS is well suited to handle this complexity. FTS also provides insights into the co-localization of ions and proteins that is consistent with NMR. By using different combinations of ions they confirm the robustness of the prediction that Caprin1 shows salt-dependent reentrant behavior, adding further support that the differential behavior of Caprin1, and pY-Caprin1 is likely to be mediated by charge-charge interactions.

We are encouraged by this reviewer’s positive assessment of our manuscript.

Reviewer #1 (Recommendations For The Authors):

Analysis:

Analyze the simulation results to distinguish net charge neutralization and interchain bridging; see MacAinsh et al.

Please see response above to points (3) and (4) under “Weaknesses” in this reviewer’s public review. We have now added a 1.5-page subsection starting from the bottom of p. 12 to the top of p. 14 to discuss a new extensive analysis of Arg+–Cl−–Arg+ configurations to identify bridging interactions, with key results reported in a new Fig. 6 (p. 42). Recent results from MacAinsh, Dey & Zhou (cited now as ref. 72) are included in the added discussion. Relevant advances made in MacAinsh et al., including clarification and classification of salt-mediated interactions in the phase separation of A1-LCD are now mentioned multiple times in the revised manuscript (p. 5, lines 19–20; p. 6, lines 2–5; p. 11, line 30; p. 14, line 10; p. 18, lines 28–29; and p. 20, line 4).

Writing and presentation

(1) Cite subtle effects that may be missed by the coarser approaches in this study

Please see response above to point (1) under “Weaknesses” in this reviewer’s public review.

(2) Try to distill the findings into a simple set of conclusions

Please see response above to point (2) under “Weaknesses” in this reviewer’s public review.

(3) Clarify and simplify physical interpretations

Please see response above to point (2) under “Weaknesses” in this reviewer’s public review.

(4) Explain the treatment of ATP-Mg as either a single ion or two separate ions; reconsider modifying the reference to ATP in the title

Please see response above to point (4) under “Weaknesses” in this reviewer’s public review.

(5) Minor points:

p. 4, citation of ref 56: this work shows ATP is a driver of LLPS, not merely a regulator (promotor or suppressor)

This citation to original ref. 56 (now ref. 63) on p. 4 is now corrected (bottom line of p. 4).

p. 7 and throughout: “using bulk [Caprin1]” – I assume this is the initial overall Caprin1 concentration. It would avoid confusion to state such concentrations as “initial” or “initial overall”

We have now added “initial overall concentration” in parentheses on p. 8 (line 4) to clarify the meaning of “bulk concentration”.

p. 7 and throughout: both mM (also uM) and mg/ml have been used as units of protein concentration and that can cause confusion. Indeed, the authors seem to have confused themselves on p. 9, where 400 (750) mM is probably 400 (750) mg/ml. The same with the use of mM and M for salt concentrations (400 mM Mg2+ but 0.1 and 1.0 M Na+)

Concentrations are now given in both molarity and mass density in Fig. 1 (p. 37), Fig. 2 (p. 38), Fig. 4 (p. 40), and Fig. 7 (p. 43), as noted in the text on p. 8 (lines 4–5). Inconsistencies and errors in quoting concentrations are now corrected (p. 10, line 18, and p. 11, line 2).

p. 7, “LCST-like”: isn’t this more like a case of a closed coexistence curve that contains both UCST and LCST?

The discussion on p. 8 around this observation from Fig. 1d is now expanded, including alluding to the theoretical possibility of a closed co-existence curve mentioned by this reviewer, as follows:

“Interestingly, the decrease in some of the condensed-phase [pY-Caprin1]s with decreasing T (orange and green symbols for ≲ 20◦C in Fig. 1d trending toward slightly lower [pY-Caprin1]) may suggest a hydrophobicity-driven lower critical solution temperature (LCST)-like reduction of LLPS propensity as temperature approaches ∼ 0◦C as in cold denaturation of globular proteins [7,23] though the hypothetical LCST is below 0◦C and therefore not experimentally accessible. If that is the case, the LLPS region would resemble those with both an UCST and a LCST [4]. As far as simple modeling is concerned, such a feature may be captured by a FH model wherein interchain contacts are favored by entropy at intermediate to low temperatures and by enthalpy at high temperatures, thus entailing a heat capacity contribution in χ(T), with [7,109,110] beyond the temperature-independent ϵh and ϵs used in Fig. 1c,d and Fig. 2. Alternatively, a reduction in overall condensed-phase concentration can also be caused by formation of heterogeneous locally organized structures with large voids at low temperatures even when interchain interactions are purely enthalpic (Fig. 4 of ref. [111]).”

p. 8 “Caprin1 can undergo LLPS without the monovalent salt (Na+) ions (LLPS regions extend to [Na+] = 0 in Fig. 2e,f”: I don’t quite understand what’s going on here. Is the effect caused by a small amount of counterion (ATP-Mg) that’s calculated according to eq 1 (with z s set to 0)?

The discussion of this result in Fig. 2e,f is now clarified as follows (p. 10, lines 8–14 in the revised manuscript):

“The corresponding rG-RPA results (Fig. 2e–h) indicate that, in the present of divalent counterions (needed for overall electric neutrality of the Caprin1 solution), Caprin1 can undergo LLPS without the monvalent salt (Na+) ions (LLPS regions extend to [Na+] = 0 in Fig. 2e,f; i.e., ρs = 0, ρc > 0 in Eq. (1)), because the configurational entropic cost of concentrating counterions in the Caprin1 condensed phase is lesser for divalent (zc = 2) than for monovalent (zc = 1) counterions as only half of the former are needed for approximate electric neutrality in the condensed phase.”

p. 9 “Despite the tendency for polymer field theories to overestimate LLPS propensity and condensed-phase concentrations”: these limitations should be mentioned earlier, along with the very high concentrations (e.g., 1200 mg/ml) in Fig. 2

This sentence (now on p. 11, lines 11–18) is now modified to clarify the intended meaning as suggested by this reviewer:

“Despite the tendency for polymer field theories to overestimate LLPS propensity and condensed-phase concentrations quantitatively because they do not account for ion condensation [99]—which can be severe for small ions with more than ±1 charge valencies as in the case of condensed [Caprin1] ≳ 120 mM in Fig. 2i–l, our present rG-RPA-predicted semi-quantitative trends are consistent with experiments indicating “

In addition, this limitation of polymer field theories is also mentioned earlier in the text on p. 6, lines 30–31.

Reviewer #2 (Recommendations For The Authors):

(1) he current version of the paper goes through many different methodologies, but how these methods complement or overlap in terms of their applicability to the problem at hand may not be so clear. This can be especially difficult for readers not well-versed in these methods. I suggest the authors summarize this somewhere in the paper.

As mentioned above in response to Reviewer #1, we have now added a subsection with heading “Overview of key observations from complementary approaches” at the beginning of the “Results” section on p. 6 (lines 18–37) and the first line of p. 7 to make our paper more accessible to readers who might not be well-versed in the various theoretical and computational techniques. A few sentences to summarize our key results are added as well to the first paragraph of “Discussion” (p. 18, lines 23–26).

(2) It wasn’t clear if the authors obtained LCST-type behavior in Figure 1d or if another phenomenon is responsible for the non-monotonic change in dense phase concentrations. At the very least, the authors should comment on the possibility of observing LCST behavior using the rG-RPA model and if modifications are needed to make the theory more appropriate for capturing LCST.

As mentioned above in response to Reviewer #1, the discussion regarding possible LCSTtype behanvior in Fig. 1d is now expanded to include two possible physical origins: (i) hydrophobicity-like temperature-dependent effective interactions, and (ii) formation of heterogeneous, more open structures in the condensed phase at low temperatures. Three additional references [109, 110, 111] (from the Dill, Chan, and Panagiotopoulos group respectively) are now included to support the expanded discussion. Again, the modified discussion is as follows:

“Interestingly, the decrease in some of the condensed-phase [pY-Caprin1]s with decreasing T (orange and green symbols for ≲ 20◦C in Fig. 1d trending toward slightly lower [pY-Caprin1]) may suggest a hydrophobicity-driven lower critical solution temperature (LCST)-like reduction of LLPS propensity as temperature approaches ∼ 0◦C as in cold denaturation of globular proteins [7,23] though the hypothetical LCST is below 0◦C and therefore not experimentally accessible. If that is the case, the LLPS region would resemble those with both an UCST and a LCST [4]. As far as simple modeling is concerned, such a feature may be captured by a FH model wherein interchain contacts are favored by entropy at intermediate to low temperatures and by enthalpy at high temperatures, thus entailing a heat capacity contribution in χ(T), with [7,109,110] beyond the temperature-independent ϵh and ϵs used in Fig. 1c,d and Fig. 2. Alternatively, a reduction in overall condensed-phase concentration can also be caused by formation of heterogeneous locally organized structures with large voids at low temperatures even when interchain interactions are purely enthalpic (Fig. 4 of ref. [111]).”

(3) In Figures 4c and 4d, ionic density profiles could be shown as a separate zoomed-in version to make it easier to see the results.

This is an excellent suggestion. Two such panels are now added to Fig. 4 (p. 40) as parts (g) and (h).

Reviewer #3 (Recommendations For The Authors):

I would suggest authors make some minor edits as noted here.

(1) Please note down the chi values that were used when fitting experimental phase diagrams with rG-RPA theory in Figure 2a,b. At present there aren’t too many such values available in the literature and reporting these would help to get an estimate of effective chi values when electrostatics is appropriately modeled using rG-RPA.

The χ(T) values and their enthalpic and entropic components ϵh and ϵs used to fit the experimental data in Fig. 1c,d are now stated in the caption for Fig. 1 (p. 37). Same fitted χ(T) values are used in Fig. 2 (p. 38) as it is now stated in the revised caption for Fig. 2. Please note that for clarity we have now changed the notation from ∆h and ∆s in our originally submitted manuscript to ϵh and ϵs in the revised text (p. 7, last line) as well as in the revised figure captions to conform to the notation in our previous works [18, 71].

(2) Authors note “monovalent positive salt ions such as Na+ can be attracted, somewhat counterintuitively, into biomolecular condensates scaffolded by positively-charged polyelectrolytic IDRs in the presence of divalent counterions”. This may be due to the fact that the divalent negative counterions present in the dense phase (as seen in the ternary phase diagrams) also recruit a small amount of Na+.

The reviewer’s comment is valid, as a physical explanation for this prediction is called for. Accordingly, the following sentence is added to p. 10, lines 27–29:

“This phenomenon arises because the positively charge monovalent salt ions are attracted to the negatively charged divalent counterions in the protein-condensed phase.”

(3) In the discussion where authors contrast the LLPS propensity of Caprin1 against FUS, TDP43, Brd4, etc, they correctly note majority of these other proteins have low net charge and possibly higher non-electrostatic interaction that can promote LLPS at room temperature even in the absence of salt. It is also worth noting if some of these proteins were forced to undergo LLPS with crowding which is sometimes typical. A quick literature search will make this clear.

A careful reading of the work in question (Krainer et al., ref. 50) does not suggest that crowders were used to promote LLPS for the proteins the authors studied. Nonetheless, the reviewer’s point regarding the potential importance of crowder effects is well taken. Accordingly, crowder effects are now mentioned briefly in the Introduction (p. 4, line 13), with three additional references on the impact of crowding on LLPS added [30–32] (from the Spruijt, Mukherjee, and Rakshit groups respectively). In this connection, to provide a broader historical context to the introductory discussion of electrostatics effects in biomolecular processes in general, two additional influential reviews (from the Honig and Zhou groups respectively) are now cited as well [15, 16].

Version 2

Abstract

Abstract

Liquid-liquid phase separation (LLPS) involving intrinsically disordered protein regions (IDRs) is a major physical mechanism for biological membraneless compartmentalization. The multifaceted electrostatic effects in these biomolecular condensates are exemplified here by experimental and theoretical investigations of the different salt- and ATP-dependent LLPSs of an IDR of messenger RNA-regulating protein Caprinl and its phosphorylated variant pY-Caprinl, exhibiting, e.g., reentrant behaviors in some instances but not others. Experimental data are rationalized by physical modeling using analytical theory, molecular dynamics, and polymer field-theoretic simulations, indicating that interchain ion bridges enhance LLPS of polyelectrolytes such as Caprinl and the high valency of ATP-magnesium is a significant factor for its colocalization with the condensed phases, as similar trends are observed for other IDRs. The electrostatic nature of these features complements ATP’s involvement in π-related interactions and as an amphiphilic hydrotrope, underscoring a general role of biomolecular condensates in modulating ion concentrations and its functional ramifications.

Introduction

Broad-based recent efforts have uncovered many intriguing features of biomolecular condensates, revealing and suggesting myriad known and potential biological functions [1–3]. These assemblies are underpinned substantially, though not exclusively, by liquid-liquid phase separation (LLPS) of intrinsically disordered regions (IDRs) as well as folded domains of proteins and nucleic acids [4, 5], while more complex equilibrium and non-equilibrium mechanisms also contribute [6–14],

Electrostatics plays major roles in biophysical and biochemical processes [15, 16]. Because of the relatively high compositions of charge residues in IDRs, electrostatics is particularly important for IDR LLPS [17, 18], which is often also facilitated by π-related interactions [19, 20], hydrophobicity, hydrogen bonding [21, 22], and is modulated by temperature [7, 23], hydrostatic pressure [4, 24], osmolytes [4], RNA [25–28], salt, pH [29], molecular crowding [30–32] and post-translational modifications (PTMs) [1, 33–35]. Multivalency underlies many aspects of IDR properties [36–40]. Here, we focus primarily on how PTM- and salt-modulated multivalent charge-charge interactions might alter IDR condensate behaviors and their possible functional ramifications. In general, electrostatic effects on IDR LLPS [17, 27, 29, 41] are dependent upon their sequence charge patterns [42–48]. Intriguingly, some IDRs undergo reentrant phase separation [4] or dissolution [49] when temperature, pressure [4], salt [50, 51], RNA [49, 52], or concentrations of small molecules such as heparin [53] is varied. Reentrance, especially when induced by salt and RNA, suggest a subtle interplay between multivalent sequence-specific charge-charge interactions and hydrophobic, non-ionic [50, 51], cation-π [49, 52], or π-π interactions.

An important modulator of biomolecular LLPS is adenosine triphosphate (ATP). As energy currency, ATP hydrolysis is utilized to synthesize or break chemical bonds and drive transport to regulate “active liquid” properties such as concentration gradients and droplet sizes [9, 54], Examples include ATP-driven assembly of stress granules [55], splitting of bacterial biomolecular condensates [56], and destabilization of nucleolar aggregates [57]. ATP can also influence biomolecular LLPS without hydrolysis, akin to other LLPS promotors or suppressors [58, 59] that are effectively ligands of the condensate scaffold [60], or through ATP’s effect on lowering free [Mg2+] [61]. Notably, as an amphiphilic hydrotrope [62] with intracellular concentrations much higher than that required for an energy source, ATP is also seen to afford an important function independent of hydrolysis by solubilizing proteins, preventing LLPS and destabilizing aggregates, as exemplified by measurements on several proteins including fused in sarcoma (FUS) [63].

Subsequent investigations indicate, however, that hydrolysis-independent [ATP] effects on biomolecular LLPS are neither invariably monotonic for a given system nor universal across different systems. For instance, ATP promotes, not suppresses, LLPS of an IgGl antibody [64], basic IDPs [65], and enhances LLPS of full-length and the C-terminal domain (CTD) of FUS at low [ATP] but prevents LLPS at high [ATP] [66]. The latter reentrant behavior has been surmised to arise from ATP binding bivalently [66, 67] or trivalently [68] to charged residues arginine (R) or lysine (K) by a combination of cation-π and electrostatic interactions, an effect also seen in the ATP-mediated LLPS of basic IDPs [65]. A similar scenario was invoked for the reentrant phase behavior of transactive response DNA-binding protein of 43 kDa (TDP-43) [69].

While π-related interactions are important for biomolecular LLPS in general [19, 20] and their interplay with electrostatics likely underlies reentrant biomolecular phase behaviors modulated by RNA [49, 52] or simple salts [50], the degree to which electrostatics alone can, in large measure, rationalize hydrolysis-independent ATP-modulated biomolecular phase reentrance has not been sufficiently appreciated. This question deserves attention. For instance, the suppression of cold-inducible RNA-binding protein condensation by ATP has been suggested to be electrostatically driven [70]. The aforementioned ATP-modulated reentrant phase behavior of FUS [66, 67] is reminiscent of the 236-residue N-terminal IDR of DEAD-box RNA helicase Ddx4’s lack of LLPS at low [NaCl] (< 15-20 mM), LLPS at higher [NaCl] [71] and decreasing LLPS propensity when [NaCl] is further increased [17, 18]. Indeed, the finding that FUS CTD (net charge per residue (NCPR) = 15/156 = 0.096) exhibits ATP-dependent reentrant phase behaviors while the N-terminal domain (NCPR = 3/267 = 0.011) does not [67] is consistent with electrostatics-based theory for the difference in salt-dependent LLPS of polyelectrolytes and polyampholytes [71] and a recent atomic simulation study of direct and indirect salt effects on LLPS [72].

With this in mind, we seek to delineate the degree to which theories focusing primarily on electrostatics can rationalize experimental ATP-related LLPS data on the 103-residue C-terminal IDR of human cytoplasmic activation/proliferation-associated protein-1 (Caprinl). Full-length Caprinl (709 amino acid residues) is a ubiquitously expressed phosphoprotein that regulates stress [73–76] and neuronal [77] granules, is necessary for normal cellular proliferation [78, 79], and may be essential for long-term memory [80]. Caprinl dysfunction leads to multiple diseases such as nasopharyngeal carcinoma [81] as well as language impairment and autism spectrum disorder [82], via, e.g., Caprinl’s modulation of the function of the fragile X mental retardation protein (FMRP) [26, 33, 77]. The C-terminal 607-709 Caprinl IDR, referred to simply as Caprinl below, is biophysically and functionally significant: It is sufficient for LLPS in vitro [33], important for assembling stress granules in the cell [73, 74], and has a substantial body of experiments [33, 83–85] for comparison with theory. Since tyrosine phosphorylations of Caprinl in vivo [86] may regulate translation in neurons [33], the Caprinl system is also useful for gaining insights into phosphorregulation of biomolecular condensates [87–89].

Recent advances in theory and computation enable modeling of sequence-specific IDR LLPS [18, 72, 88, 90–96]. Among the approaches, polymer chain models of IDRs are inherently more realistic in capturing sequence properties than models without a chain description such as patchy particle theory [58]. For chain models, all-atom simulation offer a high degree of geometric and energetic realism but its high computational cost often makes it difficult to achieve sufficient sampling and equilibration for the large system sizes that are needed for modeling biomolecular LLPS [72]. However, even coarse-grained explicit-chain simulation affords more realistic geometric and energetic representations than analytical theory, but analytical theory offers significant advantages in numerical tractability [97]. For our present purposes, the analytical rG-RPA formulation [71], which synthesizes Kuhn-lengh renormalization (renormalized Gaussian, rG) and random phase approximation (RPA) [18] to treat both high-net-charge polyelectrolytes and essentially net-neutral polyampholytes [71], is particularly well suited for Caprin1 and its phosphorylated variant pY-Caprin1. To gain deeper insights into the pertinent physical principles and to assess possible limitations of this analytical approximation, we further leverage a methodological combination of rG-RPA [71], field-theoretic simulation (FTS) [91, 98], and coarse-grained explicit-chain molecular dynamics (MD) [88, 94] to better elucidate the effects of salt, phosphorylation, and ATP on LLPS of Caprin1 and pY-Caprin1.

Results

Overview of key observations from complementary approaches

The complementary nature of our multiple methodologies allows us to focus sharply on the electrostatic aspects of hydrolysis-independent role of ATP in biomolecular condensation by comparing ATP’s effects with those of simple salt. Here, Caprin1 and pY-Caprin1 are modeled minimally as heteropolymers of charged and neutral beads in rG-RPA and FTS. ATP and ATP-Mg are modeled as simple salts (single-bead ions) in rG-RPA whereas they are modeled with more structural complexity as short charged polymers (multiple-bead chains) in FTS, though the latter models are still highly coarse-grained. Despite this modeling difference, rG-RPA and FTS both rationalize experimentally observed ATP- and NaCl-modulated reentrant LLPS of Caprin1 and a lack of a similar reentrance for pY-Caprin1 as well as a prominent colocalization of ATP with the Caprin1 condensate. Consistently, the same contrasting trends in the effect of NaCl on Caprin1 and pY-Caprin1 are also seen in our coarse-grained MD simulations, though polymer field theories tend to overestimate LLPS propensity [99]. The robustness of the theoretical trends across different modeling platforms underscores electrostatics as a significant component in the diverse roles of ATP in the context of its well-documented ability to modulate biomolecular LLPS via hydrophobic and π-related effects [63, 65, 67]. Analyses of these other nonelectrostatic effects are mostly beyond the scope of the present work but their impact are nevertheless illustrated by the Flory-Huggins interactions augmented to rG-RPA to quantitatively account for experimental data and our MD simulation of the arginine-to-lysine Caprin1 mutants. These findings are detailed below.

Physical theories of Caprin1 and phosphorylated Caprin1 LLPSs as those of polyelectrolytes and polyampholytes

The 103-residue Caprin1 is a highly charged IDR with 19 charged residues [Fig. 1a and Supporting Information (SI Appendix), Fig. S1]: 15 R, 1 K, and 3 aspartic acids (D); fraction of charged residues = 19/103 = 0.184 and NCPR = 13/103 = 0.126. With a substantial positive net charge, Caprin1’s phase behaviors are markedly different from those of polyam-pholytic IDRs with nearly zero net charge such as Ddx4 to which early sequence-specific LLPS theories were targeted [17, 18]. Instead, Caprin1 behaves like chemically synthesized polyelectrolytes [100]. In contrast, when most or all of the 7 tyrosines (Y) in the Caprin1 IDR are phosphorylated (pY), negative charges are added to produce a near-net-neutral polyampholyte. Mass spectrometry indicates that the experimental sample of highly phosphorylated Caprin1 consists mainly of a mixture of IDRs with 6 or 7 phosphorylations (SI Appendix, Fig. S2). We refer to this experimental sample as pY-Caprin1 below. For simplicity, we use only the Caprin1 IDR with 7 pYs to model the behavior of this experimental sample in our theoretical/computational formulations, partly to avoid the combinatoric complexity of sequences with 5 or 6 pYs. Accordingly, since the charge of a pY is ≈ —2 at the experimental pH = 7.4, —14 charges are added to Caprin1 for our model pY-Caprin1, resulting in a polyampholyte with a very small NCPR = —1/103 = —0.00971 (Fig. 1b). Both the experimental pY-Caprin1 (NCPR ≈ ±1/103 = ±0.00971) and model pY-Caprin1 are expected to exhibit phase properties similar to other polyampholytic IDRs.

rG-RPA+FH theory predictions rationalize different salt dependence of Caprin1 and pY-Caprin1 LLPS. (a,b) Vertical lines indicate the sequence positions (horizontal variable) of positively charged residues (blue) and negatively charged residues or phosphorylated tyrosines (red) for (a) Caprin1 and (b) pY-Caprin1. (c,d) rG-RPA+FH coexistence curves (phase diagrams, continuous curves color-coded for the NaCl concentrations indicated) agree reasonably well with experiment (dots, same color code). The grey arrows in (c,d) highlight that when [NaCl] increases, LLPS propensity increases for (c) Caprin1 but decreases for (d) pY-Caprin1. As described in our prior RPA+FH and rG-RPA+FH formulations [18, 71]. the theoretical coexistence curves shown in (c,d) are determined by fitting an effective relative permittivity εr as well as the enthalpic and entropic parts of a FH parameter χ(T) = εh/T* + εs to experimental data. For the present Caprin1 and pY-Caprin1 systems, the fitted εr = 80.5, which is remarkably close to that of bulk water (εr ≈ 78.5). The fitted (εh,εs) is (1.0,0.0) for Caprin1 and (1.0.—1.5) for pY-Caprin1. These fitted energetic parameters are equivalent [18] to ΔH ≈ —1.1 kcal mol-1 and ΔS =0.0 for forming a residue-residue contact in the Caprin1 system (c) (i.e., it is enthalpically favorable), and ΔH ≈ —1.1 kcal mol-1 and ΔS ≈ —3.0 cal mol-1K-1 for forming a residue-residue contact in the pY-Caprin1 system (d) (i.e., it is enthalpically favorable and entropically unfavorable).

While sequence-specific RPA has been applied successfully to model electrostatic effects on the LLPSs of various polyampholytic IDRs [7, 18, 44, 94, 101], RPA is less appropriate for polyelectrolytes with large NCPR [102–104] because of its treatment of polymers as ideal Gaussian chains [105]. Traditionally, theories for polyelectrolytes tackle their peculiar conformations by various renormalized blob constructs [100, 102], two-loop polymer field theory [106], modified thermodynamic perturbation theory [107], and renormalized Gaussian fluctuation (RGF) theory [99, 108], among others. As such, these formulations are mostly designed for homopolymers, making it difficult to apply directly to heteropolymeric biopolymers. In order to analyze Caprin1 and pY-Caprin1 LLPSs, we utilize rG-RPA [71], which combines Gaussian chains of effective (renormalized) Kuhn length with the key idea of RGF [43].

Phase properties predicted by rG-RPA theory for Caprin1 and pY-Caprin1 with monovalent counterions and salt are in agreement with experiment

Fig. 1c and d show that the salt- and temperature (T)-dependent phase diagrams predicted by rG-RPA with an augmented Flory-Huggins (FH) mean-field χ(T) = εh/T* + εs parameter for nonelectrostatic interactions, where εh and εs are the enthalpic and entropic contributions, respectively, and T* is reduced temperature [18, 71] (Eq. 10 of ref. [18] and “rG-RPA±FH” theory in SI Appendix), are in reasonable agreement with experiment using bulk [Caprin1] (initial overall concentration) ≈ 200 μM. (Concentrations are provided in molarity and also as mass density in Fig. 1 and subsequent figures). The rG-RPA±FH results in Fig. 1c indicate that (i) Caprin1 undergoes LLPS below 20°C with 100 mM NaCl, and that (ii) LLPS propensity, quantified by the upper critical solution temperature (UCST), increases with [NaCl]. These predictions are consistent with experimental data, including the observation that Caprin1 does not phase separate at room temperature without salt, ATP, RNA or other proteins, though Caprin1 LLPS can be triggered by adding wildtype (WT) and phosphorylated FMRP and/or RNA (overall [Caprin1] ≥ 10 μM) [33], NaCl [83], or ATP (overall [Caprin1] = 400 μM) [84]. The trend here is also in line with other theories of polyelectrolytes [108]. In contrast, rG-RPA±FH results in Fig. 1d for pY-Caprin1 shows decreasing LLPS propensity with increasing [NaCl], consistent with experimental data and the expected salt dependence of LLPS of nearly net-neutral polyampholytic IDRs such as Ddx4 [18].

Interestingly, the decrease in some of the condensed-phase [pY-Caprin1]s with decreasing T (orange and green symbols for ≤ 20°C in Fig. 1d trending toward slightly lower [pY-Caprin1]) may suggest a hydrophobicity-driven lower critical solution temperature (LCST)-like reduction of LLPS propensity as temperature approaches ~ 0°C as in cold denaturation of globular proteins [7, 23] though the hypothetical LCST is below 0°C and therefore not experimentally accessible. If that is the case, the LLPS region would resemble those with both an UCST and a LCST [4]. As far as simple modeling is concerned, such a feature may be captured by a FH model wherein interchain contacts are favored by entropy at intermediate to low temperatures and by enthalpy at high temperatures, thus entailing a heat capacity contribution in χ(T), with εh → εh(T), εs → εs (T) [7, 109, 110], beyond the temperature-independent εh and εs used in Fig. 1c,d and Fig. 2. Alternatively, a reduction in overall condensed-phase concentration can also be caused by formation of heterogeneous locally organized structures with large voids at low temperatures even when interchain interactions are purely enthalpic (Fig. 4 of ref. [111]).

rG-RPA+FH theory rationalizes [NaCl]-modulated reentrant phase behavior of Caprin1. In each salt-protein phase diagram (T = 300 K), tielines (dashed) connect coexisting phases on the boundary (magenta curve) of the cyan-shaded coexistence region. For clarity, zoomed-in views of the grey-shaded part in (a, c, e, g, i. k) are provided by the plots to the right, i.e., (b, d, f, h, j, l), respectively. The solid inclined lines in (g. h, k, l) mark the minimum counterion concentrations required for overall electric neutrality. Results are shown for monovalent cation and anion with Caprin1 (a, b) or pY-Caprin1 (c, d); or monovalent cation and divalent anion with Caprin1 (e-h); or divalent cation and tetravalent anion with Caprin1 (i-l). Cation-modulated reentrant phase behaviors is seen for a wide concentration range for Caprin1 in (a, b) but only a very narrow range of high Caprin1 concentrations in (e, f, i, j). The (εh, εs) values for computing the phase diagrams here for Caprin1 and pY-Caprin1. respectively, are the same as those used for Fig. 1c and Fig. 1d.

Salt-IDR two-dimensional phase diagrams are instrumental for exploring broader phase properties

Fig. 1c and d, though informative, are computed by a restricted rG-RPA+FH that assumes a spatially uniform [Na+]. For a more comprehensive physical picture, we now examine possible differences in salt concentration between the IDR-dilute and condensed phases by applying unrestricted rG-RPA+FH to compute two-dimensional salt-Caprin1/pY-Caprin1 phase diagrams (Fig. 2).

As stated in Materials and Methods and SI Appendix, here we define “counterions” and “salt ions”, respectively, as the small ions with charges opposite and identical in sign to that of the net charge, Q, of a given polymer. For the Caprin1/NaCl system, since Caprin1’s net charge is positive, Na+ is salt ion and Cl− is counterion. Overall electric neutrality of the system implies that the concentrations (ρ’s) of polymer (ρp), counterions (ρc), and salt ions (ρs) are related by |Q|ρP+zsρs=zcρc, (1)

where zs and zc are, respectively, the valencies of salt ions and counterions. For Caprin1 and pY-Caprin1, Q = +13 and —1, respectively, and (zs, zc) = (1,1), (1,2), and (2,4) are models for different small-ion species in the system. Specifically, in Fig. 2, we identify the zs = 1 salt ion as Na+ (Fig. 2a-f) and the zc = 1 counterion as Cl− (Fig. 2a-d), the zc = 2 counterion as (ATP-Mg)2- (Fig. 2g,h), the zs = 2 salt ion as Mg2+ and the zc = 4 counterion as ATP4- (Fig. 2i—l). As mentioned above, in the present rG-RPA formulation, (ATP-Mg)2- and ATP4- are modeled minimally as a single-bead ion. They are represented by charged polymer models with more structural complexity in the FTS models below.

Behavioral trends of rG-RPA-predicted Na+-Caprin1 two-dimensional phase diagrams are consistent with experiment

Notably, Fig. 2a,b (zs = zc = 1) predicts that Caprin1 does not phase separate without Na+, consistent with experiment, indicating that monovalent counterions alone (Cl− in this case) are insufficient for Caprin1 LLPS. When [Na+] is increased, the system starts to phase separate at a small [Na+] ≤ 0.1 M, with LLPS propensity increasing to a maximum at [Na+] ~1 M before decreasing at higher [Na+], in agreement with experiment (Fig. 3a, blue data points) and consistent with Caprin1 LLPS propensity increasing with [NaCl] from 0.1 to 0.5 M (Fig. 1c). The predicted reentrant dissolution of Caprin1 condensate at high [Na+] in Fig. 2a is consistent with measurement up to [Na+] ≈ 4.6 M indicating a significant decrease in LLPS propensity when [Na+] > 2.5 M (Fig. 3a), though the gradual decreasing trend suggests that complete dissolution of condensed droplets is not likely even when NaCl reaches its saturation concentration of ~ 6 M.

Experimental demonstration of [ATP-Mg]- and [NaCl]-modulated reentrant phase behavior for Caprin1. (a) Turbidity quantified by optical density at 600 nm (OD600, normalized by peak value) to assess Caprin1 LLPS propensity at [Caprin1] = 200 μM [for ATP-Mg dependence (red), bottom scale] or [Caprin1] = 300 μM [for NaCl dependence (blue), top scale], measured at room temperature (~ 23°C). Error bars are one standard deviations of triplicate measurements, which in most cases was smaller than the plotting symbols. The ATP-Mg dependence seen here for 200 μM Caprin1 is similar to the results for 400 μM Caprin1 (Fig. 6C of ref. [84]). (b) Microscopic images of Caprin1 and pY-Caprin1 at varying [ATP-Mg] at room temperature, showing reentrant behavior for Caprin1 but not for pY-Caprin1. Each sample contains 200 μM of either Caprin1 or pY-Caprin1, with 1% of either Caprin1-Cy5 or pY-Caprin1-Cy5 (labeled with Cyanine 5 fluorescent dye) added for visualization, in a 25 mM HEPES buffer at pH 7.4. Scale bars represent 10 μm.

The negative tieline slopes in Fig. 2a,b predict that Na+ is partially excluded from the Caprin1 condensate. This “salt partitioning” is most likely caused by Caprin1’s net positive charge and is consistent with published research on polyelectrolytes with monovalent salt [108, 112, 113]. Here, the rG-RPA predicted trend is consistent with our experiment showing significantly reduced [Na+] in the Caprin1-condensed phase compared to the Caprin1-dilute phase (Table 1), although the larger experimental reduction of [Na+] in the Caprin1 condensed droplet relative to our theoretical prediction remains to be elucidated. In contrast, for the near-neutral, very slightly negative model pY-Caprin1 (Fig. 2c,d), rG-RPA predicts LLPS at [Na+] ≈ 0, and the positive tieline slopes indicate that [Na+] is higher in the condensed than in the dilute phase. Consistent with Fig. 1d, Fig. 2c shows that pY-Caprin1 LLPS propensity always decreases with increasing [Na+].

rG-RPA-predicted salt-IDR two-dimensional phase diagrams underscore effects of counterion valency on LLPS

Interestingly, a different salt dependence of Caprin1 LLPS is predicted when the salt ion remains monovalent but the monovalent counterion Cl− is replaced by a divalent zc = 2 anion modeling (ATP-Mg)2- (as a one-bead ion) under the simplifying assumption that ATP4- and Mg2+ do not dissociate in solution. The corresponding rG-RPA results (Fig. 2e–h) indicate that, in the present of divalent counterions (needed for overall electric neutrality of the Caprin1 solution), Caprin1 can undergo LLPS without the monvalent salt (Na+) ions (LLPS regions extend to [Na+] = 0 in Fig. 2e,f; i.e., ρs = 0, ρc > 0 in Eq. (1)), because the configurational entropic cost of concentrating counterions in the Caprin1 condensed phase is lesser for divalent (zc = 2) than for monovalent (zc = 1) counterions as only half of the former are needed for approximate electric neutrality in the condensed phase.

Other predicted differences between monovalent (Fig. 2a,b) and divalent (Fig. 2e,f) counterions’ impact on Caprin1 LLPS include: (i) The maximum condensed-phase [Caprin1] at low [Na+] is lower with monovalent than with divalent counterions ([Caprin1] ~ 40 mM vs. ~ 70 mM). (ii) The [Na+] at the commencement of reentrance (i.e., at the maximum condensed-phase [Caprin1]) is much higher with monovalent than with divalent counterions ([Na+] ~ 1 M vs. ~ 0.1 M). (iii) [Na+] is depleted in the Caprin1 condensate with both monovalent and divalent counterions when overall [Na+] is high (negative tieline slopes for [Na+]≥ 2 M in Fig. 2a,e). However, for lower overall [Na+], [Na+] is slightly higher in the Caprin1 condensate with divalent but not with monovalent counterions (slightly positive tieline slopes for [Na+]≤ 2 M in Fig.2e,f). This prediction suggests that under physiological [Na+]=150~170 mM, monovalent positive salt ions such as Na+ can be attracted, somewhat counterintuitively, into biomolecular condensates scaffolded by positively-charged polyelectrolytic IDRs in the presence of divalent counterions. This phenomenon arises because the positively charge monovalent salt ions are attracted to the negatively charged divalent counterions in the protein-condensed phase.

rG-RPA is consistent with experimental [ATP-Mg]-dependent Caprin1 reentrant phase behaviors

For the zs = 2, zc = 4 case in Fig. 2i—l modeling (ATP-Mg)2- complex dissociating completely in solution into Mg2+ salt ions and ATP4- counterions (modeled as single-bead ions), rG-RPA predicts Caprin1 LLPS with ATP4- (Fig. 2k,l) in the absence of Mg2+ (the LLPS region includes the horizontal axes in Fig. 2i,j), likely because the configurational entropy loss of tetravalent counterions in the Caprin1 condensate is less than that of divalent and monovalent counterions. Tetr avalent counterions also increase the theoretical maximum condensed-phase [Caprin1] to ≥ 120 mM. At the commencement of reentrance (maximum condensed-phase [Caprin1] in Fig. 2i,j), [Mg2+] ~0.4 M, which is intermediate between the corresponding [Na+] ~ 1.0 and 0.1 M, respectively, for monovalent and divalent counterions with (zs,zc) = (1,2) and (1,1). All tieline slopes for Mg2+ and ATP4- in Fig. 2i—l are significantly positive, except in an extremely high-salt region with [Mg2+]> 8M, indicating that [(ATP-Mg)2-] is almost always substantially enhanced in the Caprin1 condensate. These observations from analytical theory will be corroborated by FTS below with the introduction of structurally more realistic models of (ATP-Mg)2-, ATP4- together with the possibility of simultaneous inclusion of Na+, Cl−, and Mg2+ in the FTS models of Caprin1/pY-Caprin1 LLPS systems. Despite the tendency for polymer field theories to overestimate LLPS propensity and condensed-phase concentrations quantitatively because they do not account for ion condensation [99]—which can be severe for small ions with more than ±1 charge valencies as in the case of condensed [Caprin1] ≥120 mM in Fig. 2i—l, our present rG-RPA-predicted semi-quantitative trends are consistent with experiments indicating [ATP-Mg]-dependent reentrant phase behavior of Caprin1 (Fig. 3a, red data points, and Fig. 3b) and that [Mg2+] as well as [ATP4-] are significantly enhanced in the Caprin1 condensate by a factor of ~ 5-60 for overall [ATP-Mg] = 3-30 mM (Table 2).

Coarse-grained MD with explicit small ions is useful for investigating subtle salt dependence in biomolecular LLPS

To gain deeper insights, we extend the widely-utilized coarse-grained explicit-chain MD model for biomolecular condensates [88, 94, 114] to include explicit small cations and anions (Materials and Methods’). ATP-mediated LLPS of short basic peptides was studied recently using all-atom simulations indicating ATP engaging in electrostatic and cation-π bridging interactions [65]. Here, we limit the small ions in our coarse-grained MD simulations of Caprin1 and pY-Caprin1 LLPS to Na+ and Cl−, focusing on the physical origins of reentrance or lack thereof as well as the effects of ariginine-to-lysine (RtoK) mutations on Caprin1. Coarse-grained models allow for the study of larger systems (IDPs of longer chain lengths and more IDPs in the system), though they cannot provide insights into more subtle structural and energetic effects as in all-atom simulations [50, 65, 72]. For computational efficiency, here we neglect solvation effects that can arise from the directional hydrogen bonds among water molecules (see, e.g., ref. [115]) by treating other aspects of the aqueous solvent implicitly as in most, though not all [97, 101] applications of the methodology [88]. Several coarse-grained interaction schemes were used in recent MD simulations of biomolecular LLPS [88, 94, 96, 116–119]. Since we are primarily interested in general principles rather than quantitative details of the phase behaviors of Caprin1 and its RtoK mutants, here we adopt the Kim-Hummer (KH) energies for pairwise amino acid interactions derived from contact statistics of folded protein structures [88], which can largely capture the experimental effects of R vs K on LLPS [94].

Explicit-ion MD rationalizes experimentally observed [NaCl]-dependent Caprin1 reentrant phase behaviors and depletion of Na+ in Caprin1 condensate

Consistent with experiment (Fig. 3) and rG-RPA (Fig. 2a-d), explicit-ion coarse-grained MD results in Fig. 4 show [NaCl]-dependent reentrant phase behavior for Caprin1 but not for pY-Caprin1 (non-monotonic and monotonic trends indicated, respectively, by the grey arrows in Fig. 4a,b). In other words, the critical temperature Tcr, which is defined as the maximum temperature (UCST) of a given phase diagram (binodal, or coexistence curve), increases then decreases with addition of NaCl for Caprin1 but Tcr always decreases with increasing [NaCl] for pY-Caprin1. Moreover, consistent with the rG-RPA-predicted tielines in Fig. 2a-d (negative slopes for Caprin1 and positive slopes for pY-Caprin1), Fig. 4e,g show that Na+ is slightly depleted in the Caprin1 condensed droplet, exhibiting the same trend as that in experiment (Fig. 3a, blue data points; and Table 1) but is enhanced in the pY-Caprin1 droplet (Fig. 4f,h). Because model temperatures in Fig. 4a,b and subsequent MD results are given in units of the MD-simulated Tcr of WT Caprin1 at [NaCl] = 0 (denoted as Tcr0 here), the Tcrs of systems with higher or lower LLPS propensities than WT Caprin1 at zero [NaCl] is characterized, respectively, by Tcr/Tcr0>1 or < 1.

Explicit-ion coarse-grained MD rationalizes [NaCl]-modulated reentrant behavior for Caprin1 and lack thereof for pY-Caprin1. (a) Simulated phase diagrams (binodal curves) of Caprin1 at different temperatures plotted in units of Tcr0 (see text). Symbols are simulated data points. Continuous curves are guides for the eye. Grey arrow indicates variation in [NaCl]. (b) Same as (a) but for pY-Caprin1. (c) A snapshot showing phase equilibrium between dilute and condensed phases of Caprin1 (brown chains) immersed in Na+ (blue) and Cl− (red) ions simulated at [NaCl] = 480 mM. (d) A similar snapshot for pY-Caprin1. (e, f) Mass density profiles, ρ(z) (in units of mg/ml), of Na+, Cl−, and (e) Caprin1 or (f) pY-Caprin1 along the elongated dimension z of the simulation box showing variations of Na+ and Cl− concentrations between the protein-dilute phase (low ρ for protein) and protein-condensed phase (high ρ for protein) at the simulation temperatures indicated, (g, h) Corresponding zoomed-in concentration profiles ρ(z) in units of mM for Na+ and Cl−. Additional mass density profiles for [NaCl]= 200 mM and 400 mM are provided in Fig. S3.

Fig. 4e,g show that [Cl−] is enhanced while [Na+] is depleted in the Caprin1 droplet. By comparison, Fig. 4f,h show that both [Cl−] and [Na+] are enhanced in the pY-Caprin1 droplet with an excess of [Na+] to balance the negatively charged pY-Caprin1 (Fig. 4h). The enhancement of [Cl−] in the Caprin1 condensed phase depicted in Fig. 4f,h is further illustrated in Fig. 5a-d by comparing the entire simulation box with a condensed droplet in the middle (Fig. 5a) with individual distributions of the Caprin1 IDR (Fig. 5b), Na+ (Fig. 5c), and Cl− (Fig. 5d). A similar trend, also attributed to charge effects, was observed in explicit-water, explicit-ion MD simulations in the presence of a preformed condensate of the N-terminal RGG domain of LAF-1 with a positive net charge [120]. For Caprin1, Fig. 5e,f suggests that, as counterion, Cl− can coordinate two positively charged R residues and thereby stabilize indirect counterion-bridged interchain contacts among polycationic Caprin1 molecules to promote LLPS, consistent with an early lattice-model analysis of generic polyelectrolytes [104] and a recent atomic simulation study of the IDP Al-LCD [72].

Counterions can stabilize Caprin1 condensed phase by favorable bridging interactions, (a) Snapshot from explicit-ion coarse-grained MD under LLPS conditions for Caprin1, showing the spatial distributions of Caprin1, Na+, and Cl− (as in Fig. 4c). The three components of the same snapshot are also shown separately in (b) Caprin1, (c) Na+, and (d) Cl−, (e) A zoomed-in view of the condensed droplet [corresponding to the green box in (a)], now with a black background and a different color scheme, (f) A further zoomed-in view of the part enclosed by the green box in (e) focusing on two interacting Caprin1 chains. A Cl− ion (pink bead indicated by the arrow) is seen interacting favorably with two arginines residues (blue beads) on the two Caprin1 chains (whose uncharged residues are colored differently by yellow or orange, lysine and aspartic acids in both chains are depicted, respectively, in magenta and red).

Explicit-ion MD offers insights into counterion-mediated interchain bridging interactions among condensed Caprin1 molecules

To assess the extent to which Cl−-mediated bridging interactions (as illustrated in Fig. 5f) contribute to condensation of polyelectrolytic IDRs, we examine the relative positions of positively charged arginine residues (Arg+) and negatively charged counterions (Cl−) of a Caprin1 solution under phase-separation conditions in which essentially all Caprin1 molecules are in the condensed phase, using 4,000 frames (MD snapshots) of an equilibrated salt-free ([NaCl] = 0) ensemble of 100 WT Caprin1 chains (net charge per chain = +13) with 1,300 Cl− counterions at T<Tcr0 as an example (Fig. 6). For simplicity, we focus on Arg+-Cl− interactions because the overwhelming majority (15/16) of the positively charged residues in Caprin1 are arginines. The computed radial distribution function, ρ(r), of Cl− around a given Arg+ exhibits a sharp peak at small r that drops to a minium at r ≈ 11 Å (Fig. 6a), indicating a strong spatial association between the oppositely charged Arg+ and Cl− as expected. Indeed, within the ensemble we analyze, 5,121,148/(4,000 × 1,300) = 98.5% of the Cl− ions are within 11 Å of an Arg+. We next enumerate putative bridging interactions involving two Arg+s on different Caprin1 chains and one Cl− (Fig. 6b) by identifying three-bead configurations in which the distance of Cl− to each of the two Arg+ is ≤ 11 Å (within the dominant small-r peak of ρ(r) in Fig. 6a), which implies that the distance between the two Arg+s is < 22 Å. In our ensemble, 4,519,387/(4,000 × 1,300) = 86.9% of the Cl− counterions are identified to be in one or more of a total of 25,112,331 such putative bridging interaction configurations. This means that, on average, each Cl− is involved in 25,112,331/4,519,387 = 5.56 configurations, and thus are coordinating sa 4 Arg+s because there are 6 (≈ 5.56) ways of pairing 4 Arg+s. Fig. 6c shows the distribution of putative bridging configurations with respect to Arg+-Arg+ distance R. Spatial distributions of Cl− in these configurations are provided in Fig. 6d,e, which are quite similiar to those of isolated Arg+-Cl−-Arg+ systems for R < 14 Å (Fig. 6f,g). Among the putative bridging configurations, we make an energetic distinction between true bridging and neutralizing (screening) configurations. Physically, a true bridging configuration may be defined by an overall favorable (< 0) sum of (i) unfavorable Coulomb potential between two Arg+ and (ii) the favorable Coulomb potential between the Cl− and one of the Arg+s that is farther away from the Cl−. By the same token, a neutralizing (screening) configuration may be defined by a corresponding overall unfavorable or neutral (≥ 0) sum of these two Coulomb potentials (i.e., the farther Arg+- Cl− distance is larger than the Arg+-Arg+ distance). In this regard, and in more general terms, Cl− ions in bridging and neutralizing interactions may be considered, respectively, as a “strong-attraction promoter” and a “weak-attraction suppressor” of LLPS [58, 59].

Counterion interactions in polyelectrolytic Caprin1. Shown distributions are averaged from 4,000 equilibrated coarse-grained MD snapshots of 100 Caprin1 chains and 1,300 Cl− counterions under phase-separation conditions ( T/Tcr0 = 160/193 = 0.829) in a 115 x 115 x 1610 Å3 simulation box in which essentially all Caprin1 chains are in a condensed droplet, (a) Radial distribution function of Cl− around a positively charged arginine residue (Arg+). (b) A zoomed-in view of Fig. 5f showcasing a putative bridging configuration with a Cl− interacting favorably with a pair of Arg+s on two different Caprin1 chains. Configurational geometry is characterized by Arg+-Arg+ distance R and the distance d of the Cl− from the line connecting the two Arg+s. (c) Distribution of putative bridging interaction configurations with respect to R. Numbers of true bridging, neutralizing, and intermediate are, respectively, in blue, green and orange, (d, e) Heat maps of two-dimensional projections of spatial distributions of Cl− around two Arg+s satisfying the putative bridging interaction conditions among the MD snapshots, (f, g) Corresponding projected distributions of isolated Arg+—Cl−-Arg+ Boltzmann-averaged systems at model temperature T. Here, P(x,d) is the total density of Cl− on a circle of radius |d| perpendicular to the heat map at horizontal position x (d, f); thus the average Cl− density at a given point (x,d) is P(x, d)/2π|d|, the patterns of which are exhibited by P(x, d)/|d| heat maps in (e, g). P(x, d) is symmetric with respect to d ↔ —d by construction, i.e., P(x,d) = P(x,—d). In each heat map, the size and (ranges of) positions of model Arg+s are indicated by blue circles; the size and the position or one of two positions (at ±d) of maximum Cl− density is indicated by a magenta circle. The MD-simulated distributions of the condensed system (d, e) are quite similar to the theory-computed isolated system (f, g) for R ≤ 14 Å, indicating that individual bridging interactions in the crowded Caprin1 condensates may be understood approximately by the electrostatics of an isolated, three-bead Arg+-Cl−-Arg+ system. For larger R. the heat maps in (f, g) and (d, e) are not as similar because some of the configurations in the isolated system (f, g) are precluded by the requirement that Arg+-Cl− distance < 11 Å for putative bridging interactions in (d, e).

In the present analysis, we group putative bridging configurations by R in bins of 2 Å (Fig. 6c). Accordingly, we may classify Cl− positions satisfying the above condition of favorable (< 0) sum of Coulomb potentials for all R values within the 2 Å range of the bin as in true bridging configurations (79.6%), those Cl− positions satisfying the above condition of unfavorable (≥ 0) sum of Coulomb potentials for all R values in the 2 Å range as in neutralizing configurations (7.4%), and those that satisfy neither as “intermediate” configurations (13.0%). Even with this more stringent criterion, ≈ 80% of putative bridging configurations are true bridging configurations. Because on average a Cl− counterion known to be involved in at least one putative bridging configuration is on average participating in ~ 5-6 such configurations, the probability that it is involved in at least one true bridging configuration is very high, at ≈ 1.0 — (0.2)5 = 99.97%. Thus, even without taking into consideration bridging interactions involving lysines, we may reasonably conclude that an overwhelming majority of ≈ 87% of Cl− counterions in the coarse-grained MD system considered are engaged in condensation-driving true bridging interactions coordinating pairs of Arg+ on different Caprin1 chains. Similar extensive Cl− and Na+ bridging interactions are observed in a recent all-atom molecular study of LLPS of short peptides under a variety of overall salt concentrations [72].

Explicit-ion MD rationalizes [NaCl]-dependent phase properties of argininelysine mutants of Caprin1

We apply our MD methodology also to four RtoK Caprin1 variants, termed 15Rtol5K, 4Rto4KN, 4Rto4KM, and 4Rto4KC (SI Appendix, Fig. S1), which involve 15 or 4 RtoK substitutions [83]. The simulated phase diagrams in Fig. 7 exhibit reentrant phase behaviors for all three 4Rto4K variants. While these results are consistent with experiments showing LLPS of these 4Rto4K variants commencing at different nonzero [NaCl]s [83], the simulated reentrant dissolution is not observed experimentally, probably because the actual [NaCl] needed is beyond the experimentally investigated or physically possible range of salt concentration. Simulated reentrant phase behaviors are also seen for 15Rtol5K; but as will be explained below, its much lower simulated UCST is consistent with no experimental LLPS for this variant [83]. Since our main focus here is on general physical principles, we do not attempt to fine-tune the MD parameters for a quantitative match between simulation and experiment. Experimentally, only WT exhibits a clear trend toward reentrant dissolution of condensed droplets (with a LLPS propensity plateau at [NaCl] ≈ 1.55-2.5 M, Fig. 3a, blue data points), whereas the LLPS of 4Rto4KM and 4Rto4Kc commences at [NaCl] ≈ 1.3 M, LLPS propensity then increases with [NaCl] (a trend consistent with the MD-predicted increasing LLPS propensity at low [NaCl]s in Fig. 7b,c), but no sign of reentrant dissolution is seen up to the maximum [NaCl] = 2 M investigated experimentally for the RtoK variants (Fig. 9B of ref. [83]). In contrast, the MD phase diagrams in Fig. 7 show a maximum LLPS propensity (highest Tcr) at [NaCl] ≈ 0.5 M. This qualitative agreement with quantitative mismatch suggests that real Caprin1 LLPS is somewhat less sensitive to small monovalent ions than that stipulated by the present MD model. This question should be tackled in future studies by considering, for example, alternate pairwise amino acid interaction energies [88, 94, 96, 116–119] and their temperature dependence [4, 23].

Explicit-ion coarse-grained MD rationalizes [NaCl]-modulated phase behavior for RtoK variants of Caprin1. Four variants studied experimentally [83] are simulated: (a) 15Rto15K, in which 15 R’s in the WT Caprin1 IDR are substituted by K, (b) 4RtoKN, (c) 4RtoKM, and (d) 4RtoKC, in which 4 R’s are substituted by K in the (b) N-terminal, (c) middle, and (d) C-terminal regions, respectively. Top panels show positions of the R (dark blue) and K (cyan) along the Caprin1 IDR sequence. Lower panels are phase diagrams in the same style as Fig. 4. The phase diagrams for WT Caprin1 from Fig. 4a are included as continuous curves with no data points in (a) for comparison.

Limitations notwithstanding, the MD-simulated trend agree largely with experiment. Predicted LLPS propensities quantified by the Tcrs in Fig. 7 follow the rank order of WT > 4Rto4KM > 4RtoKN ≈ 4Rto4KC > 15Rtol5K, which is essentially identical to that measured experimentally, viz., WT > 4Rto4KM > 4RtoKC > 4Rto4KN > 15Rtol5K (Fig. 9B of ref. [83]). In comparing theoretical and experimental LLPS, a low theoretical Tcr can practically mean no experimental LLPS when the theoretical Tcr is below the freezing temperature of the real system [18, 121]. Fig. 7a shows that even the highest Tcr for 15Rtol5K (at model [NaCl] = 480 mM) is essentially at the same level as Tcr0 for WT at [NaCl] = 0 ( Tcr/Tcr0≈1 ). This MD prediction is consistent with the combined experimental observations of no LLPS for 15Rtol5K up to at least [NaCl] = 2 M and no LLPS for WT Caprin1 at [NaCl] = 0 (Fig. 9B,C of ref. [83]).

Field-theoretic simulation (FTS) is an efficient tool for studying multiplecomponent phase properties

We next turn to modeling of Caprin1 or pY-Caprin1 LLPS modulated by both ATP-Mg and NaCl. Because tackling such many-component LLPS systems using rG-RPA or explicit-ion MD is numerically challenging, here we adopt the complementary FTS approach [122] outlined in Materials and Methods for this aspect of our investigation. FTS is based on complex Langevin dynamics [123, 124], which is related to an earlier formulation for stochastic quantization [125, 126] and has been applied extensively to polymer solutions [122, 127]. Recently, FTS has provided insights into charge-sequence-dependent LLPS of IDRs [91, 97, 98, 119, 128]. The starting point of FTS is identical to that of rG-RPA. FTS invokes no RPA and is thus advantageous over rG-RPA in this regard, though it is still limited by the lattice size used for simulation and its restricted treatment of excluded volume [98]. Here we apply the protocol detailed in refs. [97, 98].

A simple model of ATP-Mg for FTS

Going beyond the single-bead model for (ATP-Mg)2- in our analytical rG-RPA theory (Fig. 2), we now adopt a 6-bead polymeric representation of (ATP-Mg)2- (Fig. 8a) in which four negative and two positive charges serve to model ATP4- and Mg2+ respectively. Modeling (ATP-Mg)2- as a short charged polymer enables application of existing FTS formulations for multiple charge sequences to systems with IDRs and (ATP-Mg)2-. While the model in Fig. 8a does not capture structural details, its charge distribution does correspond roughly to that of the chemical structure of (ATP-Mg)2-. In developing FTS models involving IDR, (ATP-Mg)2-, and NaCl, we first assume for simplicity that (ATP-Mg)2- does not dissociate and consider systems consisting of any given overall concentrations of IDR and (ATP-Mg)2- wherein all positive and negative charges on the IDR and (ATP-Mg)2- are balanced, respectively, by Cl− and Na+ to maintain overall electric neutrality (Fig. 8a).

FTS rationalizes experimental observation of Caprin1-ATP interactions, (a) The 6-bead model for (ATP-Mg)2- and the single-bead models for monovalent salt ions used in the present FTS. (b-e) Normalized protein-protein correlation functions at three [(ATP-Mg)2-] values (b, c) and protein-ion correlation functions [Eq. (7)] at [(ATP-Mg)2-]/b-3 = 0.03 (d, e) for Caprin1 (b, d) and pY-Caprin1 (c, e), computed for Bjerrum length lB = 7b. Horizontal dashed lines are unity baselines (see text), (f) Values of position-specific integrated correlation (left vertical axis) correspond to the relative contact frequencies between individual residues (i) along the Caprin1 1DR sequence with q = (ATP-Mg)2-, Na+, or Cl− under the same conditions as (d) [Eq. (S45)] (color symbols). Included for comparison are experimental NMR volume ratios V/V0 data on site-specific Caprin1-ATP association [84]. V/V0 decreases with increased contact probability, although a precise relationship is yet to be determined. Thus, the plotted 1 — V/V0 (grey data points, right vertical scale) is expected to correlate with contact frequency.

Phase behaviors can be probed by FTS density correlation functions

LLPS of FTS systems can be monitored by correlation functions [98]. Here, we compute intra-species IDR self-correlation functions Gpp(r) (Fig. 8b,c) and inter-species crosscorrelation functions Gpq(r) between the IDR and (ATP-Mg)2- or NaCl (Fig. 8d,e) at three different overall [(ATP-Mg)2-] = 10-4b-3, 0.03b-3, and 0.5b-3, where b may be taken as the peptide virtual bond length ≈ 3.8Å (Materials and Methods). The correlation functions in Fig. 8b-e are normalized by overall densities ρp0 of the IDR and ρq0 for (ATP-Mg)2-, Na+ or Cl−, wherein density is the bead density for the given molecular species in units of b-3. LLPS of the IDR is signaled by Gpp(r)/(ρp0)2 in Fig. 8b,c dropping below the unity baseline (dashed) at large distance r because it implies a spatial region with depleted IDR below the overall concentration, which is possible only if the IDR is above the overall concentration in at least another spatial region. In other words, Gpp(r)/(ρp0)2<1 for large r indicates that IDR concentration is heterogeneous and thus the system is phase separated. For small r, Gpp(r)/(ρp0)2 is generally expected to increase because IDR chain connectivity facilitates correlation among residues local along the chain. On top of this, LLPS propensity may be quantified by Gpp(r)/(ρp0)2 for small r because a higher value indicates a higher tendency for different chains to associate and thus a higher LLPS propensity [98].

FTS rationalizes [ATP-Mg]-modulated Caprin1 reentrant phase behaviors and their colocalization in the condensed phase

[(ATP-Mg)2-]-modulated reentrance is predicted by FTS for Caprin1 but not for pY-Caprin1: When [(ATP-Mg)2-]/b-3 varies from 10-4 to 0.03 to 0.5, small-r values of the Caprin1 Gpp(r) in Fig. 8b initially increase then decrease, whereas the corresponding small-r values of the pY-Caprin1 Gpp(r) in Fig. 8c decrease monotonically, consistent with rG-RPA (Fig. 2g,h,k,l) and experiment (Fig. 3). The inter-species cross-correlations in Fig. 8d,e show further that when an IDR condensed phase is present at [(ATP-Mg)2-] = 0.03b-3 (as indicated by large-r behaviors of Gpp(r)/(ρp0)2 in Fig. 8b,c), (ATP-Mg)2- is colocalized with Caprin1 or pY-Caprin1 (high value of Gpp/ρp0ρq0 for small r) in the IDR-condensed droplet. By comparison, the variation of [Na+] and [Cl−] is much weaker. For Caprin1, Cl− is enhanced over Na+ in the Caprin1 condensed phase (small-r Gpp/ρp0ρq0 of the former larger than the latter in Fig. 8d), but the reverse is seen for pY-Caprin1 (Fig. 8e). This FTS-predicted difference, most likely arising the positive net charge on Caprin1 and the smaller negative net charge on pY-Caprin1, is consistent with the MD results in Fig. 4e-h and SI Appendix, Fig. S3.

FTS rationalizes experimentally observed residue-specific binding of Caprin1 with ATP-Mg

The propensities for (ATP-Mg)2-, Na+, and Cl− to associate with each residue i along the Caprin1 IDR (i = 1,2, …,103) in FTS are quantified by the residue-specific integrated correlation in Fig. 8f, which is the integral of the corresponding Gpq(i)(r) from r = 0 to a relative short cutoff distance r = rcontact to provide a relative contact frequency for residue i and ionic species q to be in spatial proximity (Materials and Methods and SI Appendix). Notably, the residue-position-dependent integrated correlation for (ATP-Mg)2- varies significantly, exhibiting much larger values near the N-terminal and a little before the C-terminal but weaker correlation elsewhere (Fig. 8f, red symbols). The two regions of high integrated correlation (i.e., favorable association) coincide with regions with high sequence concentration of positively charged residues. This FTS prediction is remarkably similar to the experimental NMR finding that binding between (ATP-Mg)2- and Caprin1 occurs strongly at the arginine-rich N- and C-terminal regions, as indicated by the volume ratio V/V0 data in Fig. 1C of ref. [84] that quantifies the ratio of peaks in NMR spectra in the presence and absence of trace amounts of ATP-Mn. For comparison with the FTS results, this set of experimental data is replotted as 1 — V/V0 in Fig. 8f (grey symbols, right vertical axis) to illustrate the similarity in experimental and theoretical trends because 1 — V/V0 is expected to trend with contact frequency. Corresponding FTS results for Na+ and Cl− in Fig. 8f exhibit much less residue-position-dependent variation, with Cl− displaying only slightly enhanced association in the same arginine-rich regions, and Na+ showing even less variation, presumably because the positive charges on Caprin1 are already essentially neuralized by the locally associated (ATP-Mg)2- or Cl− ions. The theory-experiment agreement in Fig. 8f regarding ATP-Caprin1 interactions indicates once again that electrostatics is an important driving force underlying many aspects of experimentally observed Caprin1-(ATP-Mg)2- association.

FTS snapshots of [ATP-Mg]-modulated reentrant phase behaviors and Caprin1- ATP-Mg colocalization

The above FTS-predicted trends are further illustrated in Fig. 9 by field snapshots. Such FTS snapshots are generally useful for visualization and heuristic understanding [91, 98, 119], including insights into subtler aspects of spatial arrangements exemplified by recent studies of subcompartmentalization entailing either co-mixing or demixing in multiple-component LLPS that are veritable by explicit-chain MD [98, 119]. Now, trends deduced from the correlation functions in Fig. 8 are buttressed by the representative snapshots in Fig. 9: As the bead density of (ATP-Mg)2- is increased from 10-4b-3 to 0.03b-3 to 0.5b-3, the spatial distribution of Caprini evolves from an initially dispersed state to a concentrated droplet to a (reentrant) dispersed state again (Fig. 9a), whereas the initial dense pY-Caprin1 droplet becomes increasingly dispersed monotonically (Fig. 9b). Colocalization of (ATP-Mg)2- with both the Caprini (Fig. 9c) and pY-Caprin1 (Fig. 9d) droplets is clearly visible at [(ATP-Mg)2-] = 0.03b-3, though the degree of colocalization is appreciably higher for Caprini than for pY-Caprin1. This is likely because the positive net charge of Caprini is more attractive to (ATP-Mg)2-. By comparison, variations in Na+ and Cl− distribution between Caprin1/pY-Caprin1 dilute and condensed phases are not so discernible in Fig. 9e-h, consistent with the small differences in the corresponding FTS correlation functions (Fig. 8d,e).

FTS rationalizes colocalization of ATP-Mg with the Caprin1 condensate. FTS snapshots are from simulations at lB = 7b (same as that for Fig. 8). Spatial distributions of real positive parts of the density fields for the protein (a, b), (ATP-Mg)2- (c, d), Na+ (e, f), and Cl− (g, h) components are shown by three snapshots each for Caprin1 (left panels) and pY-Caprin1 (right panels) at different [(ATP-Mg)2-] values as indicated. Colocalization of (ATP-Mg)2- with the Caprin1 condensed droplet is clearly seen in the [(ATP-Mg)2-]/b-3 = 0.03 panel of (c).

Robustness of general trends predicted by FTS

We have also assessed the generality of the results in Figs. 8 and 9 by considering three variations in the molecular species treated by FTS: (i) Caprini or pY-Caprin1 with only Na+ and Cl− but no (ATP-Mg)2- (SI Appendix, Fig. S4), (ii) Caprini with (ATP-Mg)2- and either Na+ or Cl− (but not both) to maintain overall charge neutrality or pY-Caprin1 with (ATP-Mg)2- and Na+ as counterion but no Cl− (SI Appendix, Fig. S5), and (iii) Caprini or pY-Caprin1 with ATP4-, Mg2+, Na+ and Cl− (SI Appendix, Fig. S6). Despite these variations in FTS models, SI Appendix Figs. S4–S6 consistently show reentrant behavior for Caprini but not pY-Caprin1 and Figs.S5 and S6 both exhibit colocalization of ATP with condensed Caprini, suggesting that these features are robust consequences of the basic electrostatics at play in Caprin1/pY-Caprin1 + ATP-Mg + NaCl systems.

Discussion

It is reassuring that, in agreement with experiment, all of our electrostatics-based theoretical approaches consistently predict salt-dependent reentrant phase behaviors for Caprini, whereas pY-Caprin1 LLPS propensity decreases monotonically with increasing salt (Figs. 2, 4, 8, and 9). This effect applies to small monovalent salts exemplified by Na+ and Cl− as well as to our electrostatics-based single- and multiple-bead models of (ATP-Mg)2- or ATP4-, with ATP exhibiting a significant colocalization with the Caprini condensed phase (Figs. 2g,h,k,l and 9c) attributable to the higher valency of (ATP-Mg)2- and ATP4- than that of monovalent ions. As mentioned above, the difference in saltdependent LLPS of Caprini and pY-Caprin1 originates largely from the polyelectrolytic nature of Caprini and the polyampholytic nature of pY-Caprin1 [71] corresponding, respectively, to the “high net charge” and “screening” classes of IDPs in a more recent analysis [72].

Related studies of electrostatic effects on biomolecular condensates

Our theoretical predictions are also largely in agreement with recent computational studies on salt concentrations in the dilute versus condensed phases [120] and salt-dependent reentrant behaviors [50] of other biomolecular condensates, including explicit-water, explicit-ion atomic simulations with preformed condensates of the N-terminal RGG domain of LAF-1 [120] and of the highly positive proline-arginine 25-repeat dipeptide PR25 [129].

A recent study examines salt-dependent reentrant LLPSs of full-length FUS (WT and G156E mutant), TDP-43, bromodomain-containing protein 4 (Brd4), sex-determining region Y-box 2 (Sox2), and annexin All [50]. Unlike the requirement of a nonzero monovalent salt concentration for Caprin1 LLPS, LLPS is observed for all these six proteins with KC1, NaCl or other salts at concentrations as low as 50 mM. Also unlike Caprin1, their protein condensates dissolve at intermediate salt then re-appear at higher salt, a phenomenon the authors rationalize by a tradeoff between decreasing favorability of cation-anion interactions and increasing favorability of cation-cation, cation-π, hydrophobic, and other interactions with increasing monovalent salt [50].

Two reasons may account for this difference. First, Caprin1 does not phase separate at low salt because it is a relatively strong polyelectrolyte (NCPR = +13/103 = +0.126). By comparison, five of the six proteins in ref. [50] are much weaker polyelectrolytes or not at all, with NCPR = +14/526 = +0.0266, +13/526 = +0.0247, −7/80 = −0.0875, 0, and +3/326 = +0.00920, respectively, for FUS (WT, mutant), TDP-43, Brd4, and All. Apparently, their weak electrostatic repulsions can be overcome by favorable nonelectrostatic interactions alone to enable LLPS.

Second, compared to Caprin1, the proteins in ref. [50] are either significantly larger (WT and mutant FUS) or significantly more hydrophobic and aromatic (the other four proteins), both properties are conducive to LLPS. For instance, although Sox2’s NCPR = +14/88 = +0.159 is higher than that of Caprin1, among Sox2’s amino acid residues, 21/88 = 23.9% are large hydrophobic or aromatic residues leucine (L), isoleucine (I), valine (V), methionine (M), phenylalanine (F), or tryptophan (W), and 17/88 = 19.3% are large aliphatic residues L, I, V, or M. This amino acid composition suggests that hydrophobic or π-related interactions in Sox2 can be sufficient to overcome electrostatic repulsion to effectuate LLPS at zero salt. In contrast, the Caprin1 IDR contains merely one L; only 10/103 = 9.7% of the residues of Caprin1 are in the L, I, V, M, F, W hydrophobic/aromatic category and only 6/103 = 5.8% are in the L, I, V, M aliphatic category. The corresponding aliphatic fractions of TDP-43, Brd4 and All, at 21/80 = 26.3%, 33/132 = 25%, and 90/326 = 27.6%, respectively, are also significantly higher than that of Caprin1.

Effects of salt on biomolecular LLPS

Effects of salts on LLPS, including partition of salt into polymer-rich phases, are of long-standing interest in polymer physics [130]. In the biomolecular condensate context, the versatile functional roles of salts are highlighted by the interplay between electrostatic and cation-π interactions [131, 132], salts’ modulating effects on heat-induced LLPSs of RNAs [133], their regulation of condensate liquidity [134], and even their potential impact in extremely high-salt exobiological environments [135]. While some of these recent studies focus primarily on salts’ electrostatic screening effects without changing the signs of the effective polymer charge-charge interaction [131], effective attractions between like charges bridged by salt or other oppositely-charged ions [104] as illustrated by Caprin1 (Fig. 5f, Fig. 6) and a recent study of Al-LCD [72] are likely needed to account for phenomena such as salt-induced dimerization of highly charged, medically relevant arginine-rich cell-penetrating short peptides [136, 137].

Tielines in protein-salt phase diagrams

In view of Caprin1’s polyelectrolytic nature, the mildly negative tieline slopes in Fig. 2a,b are consistent with rG-RPA predictions for a fully charged polyelectrolyte (Fig. 10a of ref. [71]). This depletion of monovalent salt in the condensed phase is similar to that observed in the complex coacervation of oppositely charged polyelectrolytes [138–140]. By comparison, the positive rG-RPA tieline slopes for polyampholytic pY-Caprin1 (Fig. 2c,d), confirmed by MD in Fig. 4f,h, are appreciably steeper than that predicted for fully charged (±1) diblock polyampholytes by rG-RPA and the essentially flat tielines predicted by FTS (Fig. 10b of ref. [71] and Fig. 7 of ref. [92]). Whether this difference originates from the presence of divalently charged (—2) phosphorylated sites in pY-Caprin1 remains to be elucidated. In any event, tieline analysis is generally instrumental for revealing details, such as stoichiometry, of the interactions driving multiple-component biomolecular LLPSs [11, 141], rG-RPA should be broadly useful as a computationally efficient tool for this purpose [71].

Counterion valency

Our rG-RPA prediction that the maximum condensed-phase [Caprin1] at low [Na+] is substantially higher with divalent than with monovalent counterions is in line with early findings that higher-valency counterions are more effective in bridging polyelectrolyte interactions to favor LLPS [142] and recent observations that salt ions with higher valencies enhance biomolecular LLPS [143, 144]. The possibility that this counterion/salt effect on LLPS may be exploited more generally for biological functions and/or biomedical applications remains to be further explored. In this regard, while recognizing that ATP can engage in π-related interactions [65–67], our electrostatics-based perspective of ATP-dependent reentrant phase behaviors is consistent with recent observations on polylysine LLPS modulated by enzymatically catalyzed ATP turnovers [140, 145].

Prospective extensions of the present theoretical methodology

Beyond the above comparisons, further experimental testing of other aspects of our theoretical predictions should be pursued, especially those pertaining to pY-Caprin1. Future theoretical efforts should address a broader range of scenarios by independent variations of [ATP4-], [Mg2+], [Na+], [Cl−] and to account for nonelectrostatic aspects of ATP-Mg dissociation [146] with predictions such as tieline slopes analyzed in detail to delineate effects of configurational entropy of salt ions [147] and solvent quality [148]. In addition to our basic modeling constructs, the impact of excluded volume and solvent/cosolute-mediated temperature-dependent effective interactions should be incorporated. Excluded volume is known to affect LLPS [92], demixing of IDP species in condensates [98], and partition of salt ions in polymer LLPS [139]. Moreover, LCST can be driven not only by hydrophobicity [4, 7, 23] but also by electrostatics, as suggested by experiment on complex coacervates of oppositely charged polyelectrolytes [149]. Bringing together these features into a comprehensive formulation will afford a more accurate physical picture.

Summary

To recapitulate, we have employed three complementary theoretical and computational approaches to account for the interplay between sequence pattern, phosphorylation, counterion, and salt in the phase behaviors of IDPs. Application to the Caprin1 IDR and its phosphorylated variant pY-Caprin1 provides physical rationalization for a variety of trends observed in experiments, including reentrance behaviors and very substantial ATP colocalization. These findings support a significant—albeit not exclusive—role of electrostatics in these biophysical phenomena, providing physical insights into effects of sequence-specific charge-charge interactions on ATP-modulated physiological functions of biomolecular condensates such as regulation of ion concentrations. The approach developed here should be of general utility as a computationally efficient tool for hypothesis generation, design of new experiments, exploration and testing of biophysical scenarios, as well as a starting point for more sophisticated theoretical/computational modeling.

Materials and methods

Further details of the experimental and theoretical/computational methodologies outlined below are provided in SI Appendix.

Experimental sample preparation

The low complexity 607-709 domain of Caprin1 was expressed and purified as before [33, 84]. WT Caprin1 was used in all experiments except those on [NaCl] dependence reported in Table 1 and Fig. 3a, for which a double mutant was used because residue pairs N623-G624 and N630-G631 in WT Caprin1 form isoaspartate (IsoAsp) glycine linkages over time which alters the charge distribution of the IDR [83].

Phosphorylation of the Caprin1 IDR

Phosphorylation of the WT Caprin1 IDR was performed as described in our prior study [33] by using the kinase domain of mouse Eph4A (587-896) [150] with an N-terminal His-SUMO tag.

Determination of phase diagrams

We established phase diagrams for Caprin1 and pY-Caprin1 by measuring the protein concentrations in dilute and condensed phases across a range of [NaCl]s (Fig. 1c,d). Initially homogenizing the two phases of the demixed samples into a milky dispersion through vortexing, ~ 200 μL aliquots were then incubated in a PCR thermocycler with a heated lid at 90°C, in triplicate, for a minimum of one hour. During incubation, the condensed phase settled and formed a clear phase at the bottom. For concentration measurements, the samples were diluted in 6 M GdmCl and 20 mM NaPi (pH 6.5). The dilute phase (top layer) was analyzed through a tenfold dilution of 10 μL samples, and the condensed phase (bottom layer) was analyzed through 250- to 500-fold dilution of 2 or 10 μL samples.

Concentrations of salt and ATP-Mg in dilute and condensed phases

Inductively coupled plasma optical emission spectroscopy (ICP-OES) measurements of [Na+] were performed using a Thermo Scientific iCAP Pro ICP-OES instrument in axial mode. ICP-OES was also used to determine [ATP] and [Mg2+] (Table 2). The detection of phosphorus and magnesium served as proxies for quantifying ATP and Mg2+ levels, respectively. Standard curves were prepared using solutions with known [ATP] and [Mg2+], ranging from 0 to 90 ppm for ATP and 0 to 25 ppm for Mg2+.

Caprin1 phase separation propensity at high salt concentrations

A 6 mM solution of double-mutant Caprin1 IDR (see above) in buffer (25 mM sodium phosphate, pH 7.4) was prepared by exchanging (3 times) the purified protein after size exclusion chromatography using centrifugal concentrators (3 kDa, EMD Millipore). Caprin samples for turbidity measurements were prepared by taking 0.5 μL of the above solution and diluting it into buffer (25 mM sodium phosphate, pH 7.4) containing varying [NaCl]s ranging from 0 to 4.63 M, in a sample volume of 9 μL, so as to achieve [Caprin1] of 300 μM. After rigorous mixing, 5 μL samples were loaded into a μCuvette G1.0 (Eppendorf). OD600 measurements (Fig. 3a) were recorded three times using a BioPhotometer D30 (Eppendorf).

[ATP-Mg]-dependent Caprin1 phase behaviors

Turbidity assays were conducted using the method we described previously [83].

Sequence-specific theory of heteropolymer phase separation

As detailed in refs. [97, 119], an example of the sequence-specific polymer theories [18, 71] is that for a solution with a single species of charged heteropolymers in np copies, nc counterions (same type), and ns salt ions (same type, but different from the counterions). Each polymer chain has N monomers (residues) with charge sequence |σ〉 = [σ1, σ2, …σN]T in vector notation, where σi ∈ {0, ±1, —2} is the charge of the ith residue. The counterions and salt ions are monomers carrying zc and zs charges, respectively. The particle-based partition function is given by (2)

where nw denotes the number of water (solvent) molecules, Rα,i is the position vector of the ith residue of the αth polymer, ra is the position vector of the ath small ion. accounts for polymer chain connectivity modeled by a Gaussian elasticity potential with Kuhn length l. describes the interactions among all molecular components of the system, here consisting only of Coulomb electrostatics (el) and excluded-volume (ex) for simplicity, viz., . Their interaction strengths are governed by the Bjerrum length lB and the two-body excluded volume parameter v2. By introducing conjugate fields ψ(r), w(r) and applying the Hubbard-Stratonovich transformation, the system defined by the particle-based partition function in Eq. (2) is recast as a field theory of ψ, w in which their interactions with polymer, salt, and counterion are described, respectively, by singlemolecule partition functions , , and . For instance, (3)

where is the single-polymer Hamiltonian and the chain label α is dropped.

Renormalized-Gaussian random-phase-approximation (rG-RPA)

Following refs. [43, 71], can be separated into a Gaussian-chain Hamiltonian with an effective (renormalized) Kuhn length l1 = xl and a remaining term, , where (4a) (4b)

with 12 = —1. By requiring the observable polymer square end-to-end distance be properly quantified by , x can be approximated by variational theory [43]. RPA can then be applied to the renormalized Gaussian (rG) chain system with l → xl and a corresponding scaling of the contour length to arrive at an improved theory, rG-RPA, for sequence-specific LLPS.

Explicit-ion coarse-grained molecular dynamics (MD)

The MD model in this work augments a class of implicit-water coarse-grained models [88, 94] that utilize a “slab” approach for efficient equilibration [114] by incorporating explicit small ions. As before [94], the total MD potential energy UT is the sum of long-spatial-range electrostatic (el) and short-spatial-range (sr) interactions of the Lennard-Jones (LJ) type as well as bond interactions, i.e., UT = Uel + Usr + Ubond. With small ions, the electrostatic component is given by a sum of polymer-polymer (pp), polymer-ion (pi), and ion-ion (ii) contributions: Uel = Uel,pp+Uel,pi+Uel,ii. Details of these terms are provided in SI Appendix.

Field-theoretic simulation (FTS)

FTS is useful for sequence-specific multiple-component LLPSs encountered in biomolecular settings. The new applications developed here are based on recent advances (see, e.g., refs. [91, 92, 97, 98, 122, 127]). Consider the field theoretic Hamiltonian (5)

where is single-molecule partition function [here m labels the components in the system, cf. Eq. (3)] and the breves denote convolution with Γ, i.e., for a generic field ϕ, ϕ⌣(r) = Γ*ϕ(r)≡∫dr′ Γ(r−r′)ϕ(r′) ; here ϕ = w, ψ, and Γ is a Gaussian smearing function [97]. FTS utilizes the Complex-Langevin (CL) method [123, 124] by introducing an artifical CL time variable (t), viz., w(r) → w(r, t), ψ(r) → ψ(r, t) and letting the system evolve in CL time in accordance with a collection of Langevin equations ∂ϕ(r,t)∂t=−δHδϕ(r,t)+ηϕ(r,t), ϕ=w,ψ, (6)

where the Gaussian noise ηϕ(r, t) satisfies 〈ηϕ(r,t)ηϕ′(r′,t′)〉=2δϕ,ϕ′δ(r−r′)δ(t−t′) . Thermal averages of thermodynamic observables are then computed as asymptotic CL time averages of the corresponding field operators. Spatial information about condensation and proximity of various components is readily gleaned from density-density correlation functions [97, 98], Gm,n(|r−r′|)=−〈ρ^m(r)ρ^n(r′)〉, (7)

where m, n are labels for the components in the model system. For instance, m may represent all polymer beads (denoted “p”) irrespective of the sequence positions of the beads [ρ^p(r)=∑i=1np∑α=1NΓ(r−Rα,i)] , and n may represent all six beads in our ATP-Mg model (Fig. 8a). One may also define Gpq(i)(|r−r′|)=〈ρ^p,i(r)ρ^q(r′)〉, (8)

where (i) represents the ith residue along a protein chain [ρ^p,i(r)≡∑α=1NΓ(r−Rα,i)] is the density of the ith residue among all the protein chains], and q = (ATP-Mg)2-, Na+, or Cl−. With this definition, residue-specific relative contact frequencies are estimated by integrating Eq. (8) over a spherical volume within a small inter-component distance icontact: (9)

For the normalized plotted in Fig. 8f, ρp,i0 and ρq0 are bulk (overall) densities, respectively, of the ith protein residue and of (ATP-Mg)2- or small ions, and rcontact ≈ 1-5b is used to characterize contacts. Further details are provided in SI Appendix.